---
title: "Factorial Experiment - Non-significant Interaction"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

A researcher designed a factorial experiment in a Completely Randomized Design (CRD) with 5 replications to evaluate the effect of harvest time and combine type on grain loss (kg/ha). Three harvest times (H1, H2, H3) and two types of combine harvesters (T1, T2) were tested.

```{r}
rm(list=ls())

# Load dataset
data <- read.table("DataColheitadeira.txt", header=TRUE)

# Convert to factors
data$trat <- factor(data$trat)
data$horario <- factor(data$horario)
data$tipo <- factor(data$tipo)
data$rep <- factor(data$rep)

str(data)
head(data)

attach(data)

```

### Exploratory graphs
```{r}
library(lattice)

# By harvester type
xyplot(perda ~ horario, groups = tipo, data = data, jitter.x = TRUE,
       xlab = "Harvest Time", ylab = "Loss (kg/ha)", type = c("p", "a"),
       lty = c(1, 1), lwd = c(2, 2), col = c("red", "blue"),
       key = list(x = 0.02, y = 0.9, columns = 1,
                  text = list(lab = c("Type 1", "Type 2")),
                  lines = list(lty = c(1, 1), lwd = c(2, 2), col = c("red", "blue"))))

# By harvest time
xyplot(perda ~ tipo, groups = horario, data = data, jitter.x = TRUE,
       xlab = "Harvester Type", ylab = "Loss (kg/ha)", type = c("p", "a"),
       lty = 1, lwd = 2, col = c("red", "blue", "green"),
       key = list(x = 0.02, y = 0.9, columns = 1,
                  text = list(lab = c("H1", "H2", "H3")),
                  lines = list(lty = 1, lwd = 2, col = c("red", "blue", "green"))))

```

### Assumption Checks
```{r}
fit1 <- aov(perda ~ trat)
anova(fit1)

# Normality test
shapiro.test(residuals(fit1))

# Homogeneity of variances
bartlett.test(residuals(fit1) ~ trat)
```

### ANOVA
```{r}
fit2 <- aov(perda ~ horario * tipo)
anova(fit2)
```

### Multiple comparisons
```{r}
library(agricolae)

# Extract residual degrees of freedom and mean square error
glr <- anova(fit2)[4,1]
sqr <- anova(fit2)[4,2]
qmr <- anova(fit2)[4,3]

# Tukey HSD for Harvest Time
HSD.test(y = perda, trt = horario, DFerror = glr, MSerror = qmr, alpha = 0.05)

# Tukey HSD for Harvester Type
HSD.test(y = perda, trt = tipo, DFerror = glr, MSerror = qmr, alpha = 0.05)

```

### Scheffé contrast test
```{r}
library(DescTools)

# Contrast: C1 = 2*H1 - H2 - H3
scheffe <- ScheffeTest(fit2, which = "horario", contrasts = matrix(c(2, -1, -1)), conf.level = 0.95)
scheffe
plot(scheffe)
```

### Analysis using ExpDes.pt package
```{r}
library(ExpDes.pt)

fit3 <- fat2.dic(horario, tipo, perda, 
                 quali = c(TRUE, TRUE), 
                 mcomp = "tukey", 
                 fac.names = c("Harvest Time", "Harvester Type"),
                 sigT = 0.05, sigF = 0.05)

```



