---
title: "Factorial Experiment - Randomized Complete Block Design (RCBD) - Non-Significant Interaction"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction
This is a factorial experiment arranged in a Randomized Complete Block Design (RCBD) to evaluate the effects of two factors (A and B) on the variable fat content (gordura). The interaction between factors is considered non-significant.

### Data loading and preparation
```{r}
rm(list=ls())
dados <- read.table("DataSorvete.txt", header=TRUE)
dados$trat <- factor(dados$trat)
dados$bloco <- factor(dados$bloco)
dados$A <- factor(dados$A)
dados$B <- factor(dados$B)
str(dados)
head(dados)
attach(dados)
```

### Exploratory plots
```{r}
library(lattice)

# By Factor B
xyplot(gordura ~ A, groups = B, data = dados, jitter.x = TRUE, type = c("p", "a"))

# By Factor A
xyplot(gordura ~ B, groups = A, data = dados, jitter.x = TRUE, type = c("p", "a"))
```

### Assumption checking
```{r}
fit1 <- aov(gordura ~ trat + bloco)
anova(fit1)

# Normality of residuals
shapiro.test(residuals(fit1))

# Homogeneity of variances
bartlett.test(residuals(fit1) ~ trat)
```

### Analysis of variance (Anova)
```{r}
fit2 <- aov(gordura ~ bloco + A * B)
anova(fit2)

# Mean table
model.tables(fit2, type = "means")
```

### Multiple comparisons
```{r}
library(agricolae)

# Extract residual degrees of freedom and mean square error
glr <- df.residual(fit2)
sqr <- deviance(fit2)
qmr <- sqr / glr
```

### Tukey's HSD Test
```{r}
# For Factor A
tuk_A <- HSD.test(y = gordura, trt = A, DFerror = glr, MSerror = qmr, alpha = 0.05)
tuk_A

# For Factor B
tuk_B <- HSD.test(y = gordura, trt = B, DFerror = glr, MSerror = qmr, alpha = 0.05)
tuk_B
```

### Custom Contrasts for Factor B (t-tests)
```{r}
library(gmodels)

# Define contrast matrix
cmat <- rbind("B1+B2+B3 vs B4+B5" = c(2, 2, 2, -3, -3),
              "B1 vs B2+B3"       = c(2, -1, -1, 0, 0),
              "B2 vs B3"          = c(0, 1, -1, 0, 0),
              "B4 vs B5"          = c(0, 0, 0, 1, -1))
cmat

# Apply t-tests using fit.contrast
testet <- fit.contrast(fit2, B, cmat, conf.int = 0.95)
testet

# Scheffé's test for the same contrasts
sch2 <- ScheffeTest(fit2, which = "B", contrasts = cmat, conf.level = 0.95)
sch2
```

### Using the ExpDes.pt Package
```{r}
library(ExpDes.pt)

fit3 <- fat2.dbc(A, B, bloco, gordura,
                 quali = c(TRUE, TRUE),
                 mcomp = "tukey",
                 fac.names = c("A", "B"),
                 sigT = 0.05, sigF = 0.05)
```

