---
title: "Factorial Experiment"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Factorial Experiment

## Example 1: Non-significant Interaction (CRD)

A researcher designed a factorial experiment in a Completely Randomized
Design (CRD) with 5 replications to evaluate the effect of harvest time
and combine type on grain loss (kg/ha). Three harvest times (H1, H2, H3)
and two types of combine harvesters (T1, T2) were tested.

```{r}
rm(list=ls())

data <- read.table("DataColheitadeiraNS.txt", header=TRUE)

data$trat <- factor(data$trat)
data$horario <- factor(data$horario)
data$tipo <- factor(data$tipo)
data$rep <- factor(data$rep)
```

### 1.1 Exploratory analysis

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# install.packages("lattice")
library(lattice)
```

By harvester type

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
xyplot(perda ~ horario, 
       groups = tipo, 
       data = data, jitter.x = TRUE,
       xlab = "Harvest Time", ylab = "Loss (kg/ha)", type = c("p", "a"),
       lty = c(1, 1), lwd = c(2, 2), col = c("red", "blue"),
       key = list(x = 0.02, y = 0.9, columns = 1,
                  text = list(lab = c("Type 1", "Type 2")),
                  lines = list(lty = c(1, 1), lwd = c(2, 2), col = c("red", "blue"))))
```

By harvest time

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
xyplot(perda ~ tipo, 
       groups = horario, 
       data = data, jitter.x = TRUE,
       xlab = "Harvester Type", ylab = "Loss (kg/ha)", type = c("p", "a"),
       lty = 1, lwd = 2, col = c("red", "blue", "green"),
       key = list(x = 0.02, y = 0.9, columns = 1,
                  text = list(lab = c("H1", "H2", "H3")),
                  lines = list(lty = 1, lwd = 2, col = c("red", "blue", "green"))))

```

### 1.2 Field layout

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
source("C:/StatisticsCourse/05-DIC/ExperimentalDesignFunction.R")

set.seed(7)
field_layout(data = data,
             n_row = nrow(data)/length(levels(data$rep)),
             n_col = length(levels(data$rep)),  
             design = "CRD")$plot
```

### 1.3 Assumption Checks

```{r}
fit1 <- aov(data$perda ~ data$trat)
anova(fit1)
```

**1.3.1 Checking normality**

```{r}
shapiro.test(residuals(fit1))
```

**1.3.2 Checking residual homoscedasticity**

```{r}
bartlett.test(residuals(fit1) ~ data$trat)
```

### 1.4 Analysis of variance

```{r}
fit2 <- aov(data$perda ~ data$horario * data$tipo)
anova(fit2)
```

### 1.5 Multiple comparisons

Analysis using `ExpDes.pt` package for it, we can use the `fat2.dic()`
function

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(ExpDes.pt)

fit3 <- fat2.dic(data$horario, data$tipo, data$perda, 
                 quali = c(TRUE, TRUE), 
                 mcomp = "tukey", 
                 fac.names = c("Harvest Time", "Harvester Type"),
                 sigT = 0.05, sigF = 0.05)
```

## Example 2: Non-significant Interaction (RCBD)

This is a factorial experiment arranged in a Randomized Complete Block
Design (RCBD) to evaluate the effects of two factors (A and B) on the
variable fat content (gordura). The interaction between factors is
considered non-significant.

```{r}
rm(list=ls())

data <- read.table("DataSorveteNS.txt", header=TRUE)
data$trat <- factor(data$trat)
data$bloco <- factor(data$bloco)
data$A <- factor(data$A)
data$B <- factor(data$B)
```

### 2.1 Exploratory analysis

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(lattice)

xyplot(gordura ~ A, 
       groups = B, 
       data = data, jitter.x = TRUE, type = c("p", "a"))

xyplot(gordura ~ B, 
       groups = A, 
       data = data, jitter.x = TRUE, type = c("p", "a"))
```

### 2.2 Field layout

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
source("C:/StatisticsCourse/05-DIC/ExperimentalDesignFunction.R")

set.seed(7)
field_layout(data = data,
             n_row = nrow(data)/length(levels(data$bloco)),
             n_col = length(levels(data$bloco)),  
             design = "DBC")$plot
```

### 2.3 Assumption checking

```{r}
fit1 <- aov(data$gordura ~ data$trat + data$bloco)
anova(fit1)
```

**2.3.1 Checking normality**

```{r}
shapiro.test(residuals(fit1))
```

**2.3.2 Checking residual homoscedasticity**

```{r}
bartlett.test(residuals(fit1) ~ data$trat)
```

### 2.4 Analysis of variance

```{r}
fit2 <- aov(data$gordura ~ data$bloco + data$A * data$B)
anova(fit2)

model.tables(fit2, type = "means")
```

### 2.5 Multiple comparisons

```{r}
library(ExpDes.pt)

fit3 <- fat2.dbc(data$A, data$B, data$bloco, data$gordura,
                 quali = c(TRUE, TRUE),
                 mcomp = "tukey",
                 fac.names = c("A", "B"),
                 sigT = 0.05, sigF = 0.05)
```

## Example 3: Significant Interaction (CRD)

A factorial experiment was conducted using a Completely Randomized
Design (CRD) to evaluate the effect of two factors --- Container and
Species --- on seedling height at 80 days of age.

**Container levels:**

-   Small plastic bag (R1)

-   Large plastic bag (R2)

-   Laminated bag (R3)

**Species levels:**

-   *Eucalyptus citriodora* (E1)

-   *Eucalyptus grandis* (E2)

```{r}
rm(list=ls())
data <- read.table("DataSignif.txt", header=TRUE)
data$trat <- factor(data$trat)
data$R <- factor(data$R)
data$E <- factor(data$E)
data$rep <- factor(data$rep)

knitr::kable(data)
```

### 3.1 Exploratory analysis

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(lattice)
```

By repetition within each species
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
xyplot(alt ~ R | E, 
       groups = rep, 
       data = data, 
       layout = c(2,1), jitter.x = TRUE,
       xlab = "Containers", 
       xlab.top = "Species", 
       ylab = "Height (cm)", pch = 16,
       lwd = c(2,2,2,2), col = c(1,2,3,4),
       key = list(x = 0.02, y = 0.3, columns = 1,
                  text = list(lab = c("R1","R2","R3","R4")),
                  lwd = 2, col = c(1,2,3,4)))
```

By species
```{r}
xyplot(alt ~ R, 
       groups = E, 
       data = data, 
       jitter.x = TRUE, 
       xlab = "Containers",
       ylab = "Height (cm)", 
       type = c("p", "a"),
       lty = c(1,1), lwd = c(2,2), 
       col = c("red","blue"),
       key = list(x = 0.02, y = 0.3, columns = 1,
                  text = list(lab = c("E1","E2")),
                  lines = list(lty = c(1,1), 
                               lwd = c(2,2), 
                               col = c("red","blue"))))
```

By container
```{r}
xyplot(alt ~ E, 
       groups = R, 
       data = data, 
       jitter.x = TRUE, 
       xlab = "Species",
       ylab = "Height (cm)", 
       type = c("p", "a"),
       lty = c(1,1,1), lwd = c(2,2,1), 
       col = c("red","blue","green"),
       key = list(x = 0.02, y = 0.3, columns = 1,
                  text = list(lab = c("R1","R2","R3")),
                  lines = list(lty = c(1,1,1), 
                               lwd = c(2,2,1), 
                               col = c("red","blue","green"))))

```

### 3.2 Field layout
```{r}
source("C:/StatisticsCourse/05-DIC/ExperimentalDesignFunction.R")

set.seed(7)
field_layout(data = data,
             n_row = nrow(data)/length(levels(data$rep)),
             n_col = length(levels(data$rep)),  
             design = "DIC")$plot
```

### 3.3 Assumption Checking
```{r}
fit1 <- aov(data$alt ~ data$R * data$E)
anova(fit1)
```

**3.3.1 Checking normality**
```{r}
shapiro.test(residuals(fit1))
```

**3.3.2 Checking residual homoscedasticity**
```{r}
bartlett.test(residuals(fit1) ~ data$trat)
```

### 3.4 ANOVA - Interaction effect
```{r}
fit2 <- aov(data$alt ~ data$E * data$R)
anova(fit2)

model.tables(fit2, type = "means", se = TRUE)
```

### 4.5 Multiple comparison
```{r}
library(ExpDes.pt)

fit3 <- fat2.dic(data$R, 
                 data$E, 
                 data$alt, 
                 quali = c(TRUE, TRUE), mcomp = "tukey",
                 fac.names = c("Containers", "Species"),
                 sigT = 0.01, sigF = 0.01)
```
