---
title: "Factorial Experiment - Significant Interaction"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---

#Introduction
A factorial experiment was conducted using a Completely Randomized Design (CRD) to evaluate the effect of two factors — Container and Species — on seedling height at 80 days of age.

-Container levels: Small plastic bag (R1), Large plastic bag (R2), Laminated bag (R3)

-Species levels: Eucalyptus citriodora (E1), Eucalyptus grandis (E2)
 
### Data loading and preparation
```{r}
rm(list=ls())
dados <- read.table("DataSignif.txt", header=TRUE)
dados$trat <- factor(dados$trat)
dados$R <- factor(dados$R)
dados$E <- factor(dados$E)
dados$rep <- factor(dados$rep)
str(dados)
head(dados)
attach(dados)
```

### Exploratory Plots
```{r}
library(lattice)

# By repetition within each species
xyplot(alt ~ R | E, groups = rep, data = dados, layout = c(2,1), jitter.x = TRUE,
       xlab = "Containers", xlab.top = "Species", ylab = "Height (cm)", pch = 16,
       lwd = c(2,2,2,2), col = c(1,2,3,4),
       key = list(x = 0.02, y = 0.3, columns = 1,
                  text = list(lab = c("R1","R2","R3","R4")),
                  lwd = 2, col = c(1,2,3,4)))

# By species
xyplot(alt ~ R, groups = E, data = dados, jitter.x = TRUE, xlab = "Containers",
       ylab = "Height (cm)", type = c("p", "a"),
       lty = c(1,1), lwd = c(2,2), col = c("red","blue"),
       key = list(x = 0.02, y = 0.3, columns = 1,
                  text = list(lab = c("E1","E2")),
                  lines = list(lty = c(1,1), lwd = c(2,2), col = c("red","blue"))))

# By container
xyplot(alt ~ E, groups = R, data = dados, jitter.x = TRUE, xlab = "Species",
       ylab = "Height (cm)", type = c("p", "a"),
       lty = c(1,1,1), lwd = c(2,2,1), col = c("red","blue","green"),
       key = list(x = 0.02, y = 0.3, columns = 1,
                  text = list(lab = c("R1","R2","R3")),
                  lines = list(lty = c(1,1,1), lwd = c(2,2,1), col = c("red","blue","green"))))

```

### Assumption Checking
```{r}
fit1 <- aov(alt ~ R * E)
anova(fit1)

# Normality of residuals
shapiro.test(residuals(fit1))

# Homogeneity of variances
bartlett.test(residuals(fit1) ~ trat)
```

### ANOVA - Interaction effect
```{r}
fit2 <- aov(alt ~ E * R)
anova(fit2)

# Means and standard errors
model.tables(fit2, type = "means", se = TRUE)
```

### Interaction Breakdown: Species within Containers
```{r}
fit2EdR <- aov(alt ~ R/E)
summary(fit2EdR)

# Find interaction terms
fit2EdR$assign
names(coef(fit2EdR))[fit2EdR$assign == 2]

# Get coefficient positions
grep("R1", names(coef(fit2EdR))[fit2EdR$assign == 2])
grep("R2", names(coef(fit2EdR))[fit2EdR$assign == 2])
grep("R3", names(coef(fit2EdR))[fit2EdR$assign == 2])

# Split interaction
summary(fit2EdR, split = list("R:E" = list(R1 = 1, R2 = 2, R3 = 3)))

```

### Interaction Breakdown: Containers within Species
```{r}
fit2RdE <- aov(alt ~ E/R)
summary(fit2RdE)

# Find interaction terms
fit2RdE$assign
names(coef(fit2RdE))[fit2RdE$assign == 2]

# Get coefficient positions
grep("E1", names(coef(fit2RdE))[fit2RdE$assign == 2])
grep("E2", names(coef(fit2RdE))[fit2RdE$assign == 2])

# Split interaction
summary(fit2RdE, split = list("E:R" = list(E1 = c(1,3), E2 = c(2,4))))
```

### Using ExpDes.pt Package
```{r}
library(ExpDes.pt)

fit3 <- fat2.dic(R, E, alt, quali = c(TRUE, TRUE), mcomp = "tukey",
                 fac.names = c("Containers", "Species"),
                 sigT = 0.01, sigF = 0.01)

```



