---
title: "Introduction to R"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1.Objects in R

R is an object-oriented language.

```{r}
x <- 10  # The '<-' operator works like '=' in Python.
x

x <- sqrt(2)  # Store the square root of 2 in x.
x

y <- sqrt(5)  # Store the square root of 5 in y.
y

y + x  # Sum of x and y.
```

### 2.Arithmetic Operations

R can be used for simple arithmetic expressions:

```{r}
1 + 2 + 3   # Addition
2 + 3 * 4   # Multiplication takes precedence
3 / 2 + 1   # Division precedes addition
4 * 3^3     # Power operation
```

### 3.Common Mathematical Functions

```{r}
sqrt(2)      # Square root
27^(1/3)     # Cube root
64^(1/4)     # Fourth root
sin(pi)      # Sine of pi (should be close to zero)
```

Some common mathematical functions in R:

-   `sqrt()`: Square root
-   `abs()`: Absolute value
-   `sin(), cos(), tan()`: Trigonometric functions
-   `asin(), acos(), atan()`: Inverse trigonometric functions
-   `sinh(), cosh(), tanh()`: Hyperbolic functions
-   `asinh(), acosh(), atanh()`: Inverse hyperbolic functions
-   `exp(), log()`: Exponential and natural logarithm

```{r}
log(5)    # Natural logarithm (ln)
log10(5)  # Logarithm base 10
log2(5)   # Logarithm base 2
exp(1.61) # Exponential function
```

### 4.Vectors

```{r}
x1 <- 10
x1

x2 <- c(1, 3, 6)
x2
x2[1]         # First element
x2[2]         # Second element
length(x2)    # Length of vector

is.vector(x2)
is.matrix(x2)
is.numeric(x2)
is.character(x2)
```

### 5.Creating Sequences

```{r}
x3 <- 1:10                 # Sequence from 1 to 10
x3

x4 <- seq(0, 1, by = 0.1)  # Sequence from 0 to 1 with step 0.1
x4
x4[x4 > 0.5]               # Filter values greater than 0.5
x4[x4 >= 0.5]              # Filter values greater than or equal to 0.5
x4 <= 0.5                  # Logical test
```

### 6.Replicating Values

```{r}
x5 <- seq(0, 1, length.out = 11)
x5

x6 <- rep(1, 5)              # Repeat 1 five times
x6

x7 <- rep(c(1, 2), c(3, 5))  # Repeat 1 three times and 2 five times
x7

x8 <- rep(1:3, each = 5)
x8

x9 <- rep(2:5, each = 5)
x9
```

### 7.Random Numbers

In statistics, most simulation methods need to be reproducible, even though they involve simulated data.

Then to achieve it, we can use the function `set.seed()`.

```{r}
set.seed(123)

x10 <- rnorm(10, mean = 70, sd = 10)
x10

sum(x10)      # Sum
mean(x10)     # Mean
var(x10)      # Variance
sd(x10)       # Standard deviation
min(x10)      # Minimum value
max(x10)      # Maximum value
```

The `summary()` function is one of the most commonly used functions in R, due to its usability.

```{r}
summary(x10)  # Summary statistics
summary(1:10)
```

### 8.Data Frames

Dataframes are data displayed in a table format

Columns can be `character`, `numeric`, `logical` and/or `factor`.

Each column should have the same type of data.

```{r}
d1 <- data.frame(X = 1:10,
                 Y = c(51, 54, 61, 67, 68, 75, 77, 75, 80, 82))
d1

summary(d1)
```

Using the `summary()` function we can summarize the data from the dataframe.

```{r}
names(d1)
d1$X  # Column X
d1$Y  # Column Y
```

We can use `$` (dollar sign) to acess a column by its name

```{r}
plot(d1)  # Scatter plot
title("Scatter Plot of Data Frame")
plot(d1$X, d1$Y)  # X vs Y
```

### 9.Creating Data

$$
\begin{array}{|c|c|c|c|c|}
\hline
\textbf{Rep} & \textbf{A} & \textbf{B} & \textbf{C} & \textbf{D} \\
\hline
1 & 25 & 31 & 22 & 33 \\
2 & 26 & 25 & 26 & 29 \\
3 & 20 & 28 & 28 & 31 \\
4 & 23 & 27 & 25 & 34 \\
5 & 21 & 24 & 29 & 28 \\
\hline
\end{array}
$$

```{r}
treatments <- rep(c("A", "B", "C", "D"), each = 5)

replicates <- rep(1:5, times = 4)

production <- c(25, 26, 20, 23, 21, 
                31, 25, 28, 27, 24, 
                22, 26, 28, 25, 29, 
                33, 29, 31, 34, 28)

data <- data.frame(treatments, replicates, production)
```

### 10. Reading Data

External data can be read from `.txt` files (plain text) or `.csv` files (comma-separated values), using respectively `read.txt()` and `read.csv()` functions.

Ensure that the file `"data_1.csv"` is present in the working directory before attempting to load it.

```{r}
data <- read.csv("data_1.csv", header = TRUE)
head(data)
```

As we can se the theatments is caracterize as `chr`, we have to change it to `factor`

```{r}
data$treatments <- as.factor(data$treatments)
head(data)
```

## 11. Others utilities

Every function in R has an underlying implementation and was developed by someone.

To access and better understand how each function works, you can consult its manual by typing `?function_name` in the R console.

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
y <- c(72.5, 69, 75, 70.8, 71.2, 73, 70, 67.1, 71, 72.5)

?sum
?mean

sum(y)
length(y)
mean(y)
median(y)
sort(y)
table(y)
var(y)
sd(y)
sqrt(var(y))
summary(y)
epm <- sd(y) / sqrt(length(y))
epm

q1 <- quantile(y, 0.25)
q2.5 <- quantile(y, 0.025)
q97.5 <- quantile(y, 0.975)
IQR(y)
boxplot(y)
```
