---
title: "Mixed Models"
author: 
  - name: PhD Student Hernandes Peres
    email: hp.panichi@ufl.edu
    affiliation:
      - name: Sweet Corn and Potato Genomics Lab. 
        city: Gainesville
        region: FL
        country: US
format: 
  html:
    toc: true
    toc-depth: 3
    toc-title: Sections
    toc-location: left
    number-sections: false
    number-depth: 4
    toc_float: true
    theme: [zephyr, theme.scss]
    embed-resources: true
    self-contained: true
    anchor-sections: true
    smooth-scroll: true
    link-external-newwindow: false
    code-fold: false
    code-overflow: wrap
    code-block-border-left: "#781c1e"
    code-copy: hover
    code-summary: "Show/Hide Code"
    lang: en
    fig-format: png
    fig-responsive: true
    fig-cap-location: bottom
    tbl-cap-location: top
    page-layout: article
    title-block-banner-color: 'white'
    highlight-style: github
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(emmeans)
library(lme4)
library(lmerTest)
library(ggplot2)
library(dplyr)
```

## Mixed Models

```{r}
rm(list = ls())
setwd("/Users/hp.panichi/OneDrive - University of Florida/My Drive/StatisticCourse/13-MixedModels/")
data <- read.table("DataMilho.txt", header = T)

data$rep <- as.factor(data$rep)
data$inc.bloco <- as.factor(data$inc.bloco)
data$gen <- as.factor(data$gen)
data$row <- data$row
data$col <- data$col

table(data$rep, data$gen)
```

### When Mixed Models takes advantages?


- **Real-world complexity:** These models acknowledge that observations within the same group are likely correlated, rather than treating every observation as completely independent;

Through utilizing the package `lme4`, and its functions `lmer()` we can model the factors as we want. To define random variables, the term most be follow by the structure `(1|variable)`


### Writing the model

$$
Y_{ij} = \mu + B_j + G_i + \varepsilon_ij
$$
- with $G_i \sim N(0, \sigma_g^2)$;

- with $\varepsilon_{ij}\sim N(0, \sigma_g^2)$;

For Mixed Models, random terms are tested for:

- $H_0: \sigma_g^2 = 0$
- $H_a: \sigma_g^2 \neq 0$

```{r}
mm = lmer(yield~inc.bloco+(1|gen), data = data)

summary(mm)
anova(mm)
```

As we can verify non significance was observed for the inc.block effect

For random terms we have to follow with the **Likelihood Ratio Test**, computed following the equation:

$$
LRT = -2 \times ln(\frac{L(Reduced)}{L(full)})
$$

- LRT should be used for nested models
- LRT Requires the fit of another model, if by removing the random term there is decrease in goodness of fit, then we have evidences that the variance is greater than zero
- To compare non-nested models, other criteria should be used, such as the AIC and BIC
- To compare fixed terms, Wald Test can be performed

```{r}
rand(mm)
```

### Shrikage term

For unbalance experimental data Mixed Models becomes important due to good statisticals properties. 

$$
\lambda = \frac{r\sigma_g^2}{r\sigma_g^2+\sigma_e^2}
$$

where the Best Linear Unbiased Predictors (**BLUPs**) are computed as follow:

$$
g_i = (\mu_i-\hat{\mu_i})\lambda
$$
And therefore, the shrinkage of an individual becomes a functions of its $\sigma_g^2$ $\sigma_e^2$ and the number of replicates ($r$).
