---
title: "Randomized Block Design"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---

## Example 1

```{r}
rm(list = ls())
data <- read.table("DataRBD.txt", header = TRUE)
data$trat <- as.factor(data$trat)
data$bloco <- as.factor(data$bloco)

n <- length(prod)
```

### 1.1 Field layout

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
source("C:/StatisticsCourse/05-DIC/ExperimentalDesignFunction.R")

set.seed(7)
field_layout(data = data,
             n_row = nrow(data)/length(levels(data$bloco)),
             n_col = length(levels(data$bloco)),  
             design = "DBC")$plot
```

### 1.2 Basic Descriptive Statistics

```{r}
colnames(data)[colnames(data) == "rep"] <- "bloco"

mean.prodt <- tapply(data$prod, data$trat, mean)
var.prodt <- tapply(data$prod, data$trat, var)
mean.prodb <- tapply(data$prod, data$bloco, mean)
var.prodb <- tapply(data$prod, data$bloco, var)
```

### 1.3 ANOVA and Residuals

```{r}
fit1 <- aov(data$prod ~ data$bloco + data$trat)
res <- residuals(fit1)
```

Residual Diagnostics

**3.1. Checking residuals normality**

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
xb <- mean(res)
sx <- sd(res)

library(car)

qqPlot(res, 
       main = "Q-Q Plot with Confidence Envelope",
       col = "blue",
       pch = 20,
       envelope = 0.95)  # 95% confidence envelope
```

```{r message=FALSE, warning=FALSE, paged.print=TRUE}
library(nortest)
tests <- list(
  ks.test(res, "pnorm", xb, sx),
  lillie.test(res),
  cvm.test(res),
  shapiro.test(res),
  sf.test(res),
  ad.test(res))

results <- data.frame(Statistic = sapply(tests, function(x) x$statistic),
                      p_value = sapply(tests, function(x) x$p.value),
                      row.names = sapply(tests, function(x) x$method))

knitr::kable(results)
```

**3.2. Checking residual Homoscedasticity**

```{r}
bartlett.test(res, data$trat)
```

**3.3. Checking residuals independence**

```{r}
plot(res,
     type = "b",
     pch = 20,
     col = "blue",
     xlab = "Observation Order",
     ylab = "Residuals",
     main = "Residuals in Order of Collection")
abline(h = 0, col = "red", lty = 2)
```

### 1.4 ANOVA Summary

Utilizing the `ExpDes.pt` package we can run ANOVA for a Randomized Completed Blocks Design, using `dbc()` function.

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(ExpDes.pt)
fit3 <- dbc(data$trat, data$bloco, data$prod, quali = TRUE, 
            mcomp = "duncan", sigT = 0.05, sigF = 0.05)
```

As we can se the F test from the ANOVA is not significant, then we should not follow with the multiple comparison

## Example 2

Cleaning environment and loading data

```{r message=FALSE, warning=FALSE}
rm(list=ls())

data <- read.table("DataSoja.txt", header=TRUE)
data$trat <- factor(data$trat)
data$bloco <- factor(data$bloco)

knitr::kable(data)
```

### 2.1 Field layout

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
source("C:/StatisticsCourse/05-DIC/ExperimentalDesignFunction.R")

set.seed(7)
field_layout(data = data,
             n_row = nrow(data)/length(levels(data$bloco)),
             n_col = length(levels(data$bloco)),  
             design = "DBC")$plot
```

### 2.2 ANOVA and residuals

```{r}
fit_list <- list(altpvagem = aov(data$altpvagem ~ data$bloco + data$trat),
                 altplanta = aov(data$altplanta ~ data$bloco + data$trat),
                 estande = aov(data$estande ~ data$bloco + data$trat),
                 prod_ton = aov(data$prod_ton ~ data$bloco + data$trat))

residuals_list <- lapply(fit_list, residuals)
lapply(fit_list, anova)
```

### 2.3 Coefficient of Variation (CV)

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(agricolae)
lapply(fit_list, cv.model)
```

Normality Test (Lilliefors)

```{r message=FALSE, warning=FALSE}
library(nortest)
lillie_results <- lapply(residuals_list, lillie.test)

stats <- sapply(lillie_results, function(x) x$statistic)
pvals <- sapply(lillie_results, function(x) x$p.value)

normality_table <- data.frame(Statistic = stats, `p-value` = pvals)

knitr::kable(normality_table)
```

### 2.4 Multivariate ANOVA

```{r}
manova_formula <- as.formula(paste0("cbind(", 
                                    paste(names(data)[-c(1,2)],
                                    collapse=","), ") ~ trat+bloco"))

fit_manova <- aov(manova_formula, data=data)

summary(fit_manova)
```

Using ExpDes.pt for RCBD analysis

```{r}
library(ExpDes.pt)
for (i in 3:ncol(data)) {
  cat("\nVariable:", names(data)[i], "\n")
  dbc(data$trat, 
      data$bloco, 
      data[,i], 
      quali=TRUE, mcomp="sk", sigT=0.05, sigF=0.05)}
```
