---
title: "Randomized Block Design"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---
## Load Required Packages

```{r}
library(nortest)
library(SuppDists)
library(outliers)
library(agricolae)
library(ExpDes.pt)
```

## Load Data and Prepare Variables
```{r}
rm(list = ls())
dados <- read.table("DataRBD.txt", header = TRUE)
dados$trat <- factor(dados$trat)
dados$bloco <- factor(dados$bloco)
attach(dados)
n <- length(prod)
```

## Basic Descriptive Statistics
```{r}
mean.prodt <- tapply(prod, trat, mean)
var.prodt <- tapply(prod, trat, var)
mean.prodb <- tapply(prod, bloco, mean)
var.prodb <- tapply(prod, bloco, var)
```

## ANOVA and Residuals
```{r}
fit1 <- aov(prod ~ bloco + trat)
res <- residuals(fit1)
CV <- (sqrt(mean(anova(fit1)[["Mean Sq"]][3])) / mean(prod)) * 100
```

## Residual Diagnostics - Normality
```{r}
xb <- mean(res)
sx <- sd(res)

# QQ Plot
qqnorm(res, col = "blue", pch = 20)
qqline(res, col = "red", lwd = 2, lty = 2)

# QQ Envelope
nsim <- 500
conf <- 0.95
dadossim <- matrix(rnorm(n * nsim, mean = xb, sd = sx), nrow = n)
dadossim <- apply(dadossim, 2, sort)
infsup <- apply(dadossim, 1, quantile, probs = c((1 - conf)/2, (1 + conf)/2))
xbsim <- rowMeans(dadossim)
faixay <- range(res, dadossim)
qq0 <- qqnorm(res, col = "blue", pch = 20, ylim = faixay)
eixox <- sort(qq0$x)
lines(eixox, xbsim, lty = 2, col = "red", lwd = 2)
lines(eixox, infsup[1, ])
lines(eixox, infsup[2, ])
```

Normality Tests
```{r}
tests <- list(
  ks.test(res, "pnorm", xb, sx),
  lillie.test(res),
  cvm.test(res),
  shapiro.test(res),
  sf.test(res),
  ad.test(res)
)

results <- data.frame(
  Statistic = sapply(tests, function(x) x$statistic),
  p_value = sapply(tests, function(x) x$p.value),
  row.names = sapply(tests, function(x) x$method)
)
print(results, digits = 4)
```

## Residual Diagnostics - Homoscedasticity

```{r}
plot(fitted(fit1), res, pch = 19, col = 4, xlab = "Fitted Values", ylab = "Residuals")
plot(trat, res, pch = 19, col = 4, xlab = "Treatments", ylab = "Residuals")
boxplot(res ~ trat, col = 2:6, xlab = "Treatments", ylab = "Residuals")
points(tapply(res, trat, mean), pch = 20, col = 1, cex = 1.5)
points(tapply(res, trat, mean) + tapply(res, trat, sd), pch = "*", col = 3)
points(tapply(res, trat, mean) - tapply(res, trat, sd), pch = "*", col = 3)
```

Homogeneity Tests
```{r}
var.res <- tapply(res, trat, var)
Fmax <- max(var.res) / min(var.res)
p.value.Fmax <- pmaxFratio(Fmax, 2, 5, lower.tail = FALSE)

cochran.test(res ~ trat, data.frame(res, trat), inlying = FALSE)
bartlett.test(res, trat)
```

## Residual Diagnostics - Independence

```{r}
plot(fitted(fit1), res, pch = 19, col = 4, xlab = "Fitted Values", ylab = "Residuals")

# Simulated order of data collection
coleta <- 1:n
plot(sort(coleta), res, type = "o", pch = 19, col = 2, xlab = "Collection Order", ylab = "Residuals")
abline(h = 0, col = "blue")
abline(h = seq(-3, 4, 1), v = seq(1, n, 1), lty = 3, col = "gray")
```

## ANOVA Summary and Post-hoc Tests

```{r}
fit2 <- aov(prod ~ bloco + trat)
anova(fit2)
duncan.test(fit2, "trat", group = TRUE, console = TRUE)
```

Using ExpDes.pt
```{r}
fit3 <- dbc(trat, bloco, prod, quali = TRUE, mcomp = "lsd", sigT = 0.05, sigF = 0.05)
```

## Exemple 3
```{r}
rm(list=ls())
```

```{r}
# Load data
data <- read.table("DataSoja.txt", header=TRUE)
data$trat <- factor(data$trat)
data$bloco <- factor(data$bloco)
attach(data)
```

ANOVA models and residuals
```{r}
fit_list <- list(
  altpvagem = aov(altpvagem ~ bloco + trat),
  altplanta = aov(altplanta ~ bloco + trat),
  estande = aov(estande ~ bloco + trat),
  prod_ton = aov(prod_ton ~ bloco + trat)
)

residuals_list <- lapply(fit_list, residuals)
lapply(fit_list, anova)
```

Coefficient of Variation (CV)
```{r}
library(agricolae)
lapply(fit_list, cv.model)
```

Normality Test (Lilliefors)
```{r}
library(nortest)
lillie_results <- lapply(residuals_list, lillie.test)

# Summary table
stats <- sapply(lillie_results, function(x) x$statistic)
pvals <- sapply(lillie_results, function(x) x$p.value)
normality_table <- data.frame(Statistic = stats, `p-value` = pvals)
print(round(normality_table, 4))
```

# Multivariate ANOVA (MANOVA)
```{r}
manova_formula <- as.formula(paste0("cbind(", paste(names(data)[-c(1,2)], collapse=","), ") ~ trat + bloco"))
fit_manova <- aov(manova_formula, data=data)
summary(fit_manova)
```

Using ExpDes.pt for RCBD analysis
```{r}
library(ExpDes.pt)
for (i in 3:ncol(data)) {
  cat("\nVariable:", names(data)[i], "\n")
  dbc(trat, bloco, data[,i], quali=TRUE, mcomp="sk", sigT=0.05, sigF=0.05)
}
```

Note: multiple comparison options include: "tukey", "duncan", "snk", "lsd", "sk"