---
title: "Correlation"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---
# 1
```{r}
# Clear environment
rm(list = ls())
ls()

# Load dataset
data <- read.table("3EST220-DataTemperatura.txt", header = TRUE)
str(data)
head(data)
attach(data)
```

Descriptive statistics

```{r}
# Variances
var(y1)
var(y2)

# Covariance
cov(y1, y2)

# Remove first column if it's not a response variable
data_clean <- data[, -1]
data_clean
```

Correlation analysis

```{r}
# Pearson correlation matrix
cor(data_clean, method = "pearson")

# Hypothesis test for correlation
cor.test(y1, y2, method = "pearson", alternative = "two.sided")
```

**Interpretation:** -H0 (null hypothesis): ρ = 0 (no correlation) -H1 (alternative hypothesis): ρ ≠ 0 -The *p-value* indicates whether the correlation is statistically significant.

Scatterplot
```{r}
# Plotting the relationship
plot(y1, y2, pch = 19, col = "blue",
     xlab = "y1", ylab = "y2", main = "Scatterplot of y1 vs y2")
```

# 2
```{r}
# Clear all objects
rm(list = ls())
ls()

# Load dataset
data <- read.table("5EST220-DataPesoAlt.txt", header = TRUE)
str(data)
head(data)
attach(data)
```

Descriptive statistics
```{r}
# Variance of each variable
var(peso)    # Variance of weight
var(altura)  # Variance of height

# Covariance between height and weight
cov(peso, altura)

# Remove the first column if it's not a response variable
data_clean <- data[, -1]
data_clean
```

Pearson correlation
```{r}
# Pearson correlation matrix
cor(data_clean, method = "pearson")

# Hypothesis test for Pearson correlation
cor.test(peso, altura, method = "pearson", alternative = "two.sided")
```

**Hypothesis Interpretation:**
H₀ (null hypothesis): ρ = 0 (no linear correlation)
H₁ (alternative hypothesis): ρ ≠ 0 (there is linear correlation)
The *p-value* will determine whether the correlation is statistically significant.

```{r}
# Scatterplot of Height vs Weight
plot(altura, peso,
     pch = 19,
     col = "blue",
     xlab = "Height (cm)",
     ylab = "Weight (kg)",
     main = "Scatterplot: Height vs Weight")
```

# 3

```{r}
# Load data
data <- read.table("7EST220-DataFeijao.txt", header = TRUE)

# Convert genotypes and blocks to factors
data$gen <- factor(data$gen)
data$bloco <- factor(data$bloco)

str(data)
head(data)

attach(data)
```

ANOVA for all response Variables
```{r}
# Load ExpDes.pt package
require(ExpDes.pt)

# Column index where response variables start
col_start <- 3

# Perform ANOVA with Scott-Knott for each response variable
for(i in col_start:ncol(data)) {
  cat("\n", "Response Variable:", names(data)[i], "\n")
  dbc(gen, bloco, data[, i], quali = TRUE, mcomp = "sk", sigT = 0.05, sigF = 0.05)
}
```

Correlation between two response variables
```{r}
# Correlation between productivity and 100-seed weight
cor(prod, pcs)

# Pearson correlation test
cor.test(prod, pcs, method = "pearson", alternative = "two.sided", conf.level = 0.95)

```

Correlation matrix for all response variables
```{r}
# Remove genotype and block columns
responses <- data[, -c(1, 2)]
head(responses)

# Correlation matrix
r <- cor(responses)
r
```

p-values for Pearson Correlation Tests

```{r}
# Matrix of p-values for correlation significance
pvalue <- matrix(0, ncol(responses), ncol(responses))
colnames(pvalue) <- rownames(pvalue) <- colnames(responses)

for(i in 1:ncol(responses)) {
  for(j in 1:ncol(responses)) {
    pvalue[i, j] <- cor.test(responses[, i], responses[, j], method = "pearson")$p.value
  }
}

pvalue
```

Correlation Between 'flor' and Other Variables

```{r}
# Correlation of 'flor' with all other variables
x <- responses[[1]]  # 'flor'
lapply(responses[-1], cor.test, x = x, method = "pearson")

```

