---
title: "Correlation"
author: 
  - name: Undergrad Hernandes Peres
    email: Hernandes.panichi@ufv.br
    affiliation:
      - name: Biometrics and Quantitative genetics Lab. 
        url: https://idata.ufv.br/
        city: Viçosa
        region: Minas Gerais
        country: Brazil
format: 
  html:
    toc: true
    toc-depth: 3
    toc-title: Sections
    toc-location: left 
    number-sections: true
    number-depth: 4
    toc_float: true
    theme: [zephyr, theme.scss]
    embed-resources: true
    self-contained: true
    anchor-sections: true
    smooth-scroll: true
    link-external-newwindow: false
    code-fold: false
    code-overflow: wrap
    code-block-border-left: "#781c1e"
    code-copy: hover
    code-summary: "Show/Hide Code"
    lang: en
    fig-format: png
    fig-responsive: true
    fig-cap-location: bottom
    tbl-cap-location: top
    page-layout: article
    csl: apa.csl
    css: styles.css
    title-block-banner: true
    title-block-banner-color: 'white'
    highlight-style: github
editor_options: 
  chunk_output_type: console
---
# Correlation

## Example 1:
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(dplyr)
library(knitr)
library(kableExtra)

rm(list = ls())
data <- read.table("DataTemperatura.txt", header = TRUE)
```

Descriptive statistics
```{r}
var(data$y1)
var(data$y2)

cov(data$y1, data$y2)

data_clean <- data[, -1]

kable(data_clean, format = "html", 
      caption = "Tabela formatada") %>%
   kable_styling(bootstrap_options = c("striped", 
                                      "hover", "condensed", "responsive"))
```

Pearson correlation matrix
```{r}
knitr::kable(cor(data_clean, method = "pearson"))
```

Hypothesis test for correlation
```{r}
cor.test(data$y1, data$y2, 
         method = "pearson", 
         alternative = "two.sided")
```

**Interpretation:** 
-$H0$ (null hypothesis): ρ = 0 (no correlation) 
-$H1$ (alternative hypothesis): ρ ≠ 0 -The *p-value* indicates whether the correlation is statistically significant.

Scatterplot
```{r}
plot(data$y1, data$y2, pch = 19, col = "blue",
     xlab = "y1", ylab = "y2", main = "Scatterplot of y1 vs y2")
```

## Example 2:
```{r}
rm(list = ls())

data <- read.table("DataPesoAlt.txt", header = TRUE)

knitr::kable(data)
```

### 2.1 Descriptive statistics
```{r}
cov(data$peso, data$altura)

data_clean <- data[, -1]
```

Pearson correlation matrix
```{r}
knitr::kable(cor(data_clean, method = "pearson"))
```

Hypothesis test for Pearson correlation
```{r}
cor.test(data$peso, data$altura, 
         method = "pearson", alternative = "two.sided")
```

**Hypothesis Interpretation:**
-   H₀ (null hypothesis): ρ = 0 (no linear correlation)
-   H₁ (alternative hypothesis): ρ ≠ 0 (there is linear correlation)

The *p-value* will determine whether the correlation is statistically significant.

```{r}
plot(data$altura, data$peso,
     pch = 19,
     col = "blue",
     xlab = "Height (cm)",
     ylab = "Weight (kg)",
     main = "Scatterplot: Height vs Weight")
```

## Example 3:
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
remove(list = ls())

require(ExpDes.pt)

data <- read.table("DataFeijao.txt", header = TRUE)

data$gen <- factor(data$gen)
data$bloco <- factor(data$bloco)
```

ANOVA for all response Variables

```{r}
col_start <- 3 # Column index where response variables start

for(i in col_start:ncol(data)) {
  cat("\n", "Response Variable:", names(data)[i], "\n")
  knitr::kable(dbc(data$gen, data$bloco, data[, i], quali = TRUE, mcomp = "sk", sigT = 0.05, sigF = 0.05))}
```

Correlation between productivity and 100-seed weight
```{r}
cor(data$prod, data$pcs)

cor.test(data$prod, data$pcs, 
         method = "pearson", 
         alternative = "two.sided", conf.level = 0.95)
```

Correlation matrix for all response variables
```{r}
responses <- data[, -c(1, 2)] # Remove genotype and block columns
knitr::kable(responses)
```


```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(ggplot2)
library(reshape2)
r <- cor(responses)
knitr::kable(r)

r <- melt(r)

r %>% ggplot(aes(x = Var1, y = Var2, fill = value))+
  geom_tile()+
  geom_text(aes(label = round(value,2)), size = 4, 
            fontface = "bold")+
  scale_fill_gradient2(low = "#d7191c",
                      mid = "#ffffbf",
                      high = "#1a9641",
                      midpoint = 0)+
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, face = "bold", hjust = 0.5),
        axis.text.y = element_text(size = 10, face = "bold", 
                                   hjust = 0.5, angle = 90),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(), 
        legend.position = "none")
```

p-values for Pearson Correlation Tests
```{r}
pvalue <- matrix(0, ncol(responses), ncol(responses))
colnames(pvalue) <- rownames(pvalue) <- colnames(responses)

for(i in 1:ncol(responses)) {
  for(j in 1:ncol(responses)) {
    pvalue[i, j] <- cor.test(responses[, i], responses[, j], method = "pearson")$p.value
  }
}

knitr::kable(pvalue)

pvalue <- melt(pvalue)
```
```{r}
pvalue %>% ggplot(aes(x = Var1, y = Var2, fill = value))+
  geom_tile()+
  geom_text(aes(label = round(value,2)), size = 4, 
            fontface = "bold")+
  scale_fill_gradient2(low = "#d7191c",
                      mid = "#ffffbf",
                      high = "#1a9641",
                      midpoint = 0.5)+
  theme_minimal()+
  theme(axis.text.x = element_text(size = 10, face = "bold", hjust = 0.5),
        axis.text.y = element_text(size = 10, face = "bold", 
                                   hjust = 0.5, angle = 90),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(), 
        legend.position = "none")
```

