---
title: "Contrast"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---

### 1. Defining data
cleaning the environment
```{r}
rm(list = ls()) 
```

Assuming 4 treatments, 6 replicates each
```{r}
trat <- gl(4, 6, labels=LETTERS[1:4]); trat
```

replicates
```{r}
r    <- rep(1:6, 4); r
```

Production data
```{r}
prod <- c(58, 49, 51, 56, 50, 48,
          60, 55, 66, 61, 54, 61,
          59, 47, 44, 49, 62, 60,
          45, 33, 34, 48, 42, 44); prod
```

Combine into a data frame
```{r}
dados <- data.frame(trat,r,prod); dados
```

Calculating treatment means
```{r}
medtrat<-tapply(prod, trat, mean);medtrat
```

### 2. Analysis of variance (ANOVA)

Check data frame and factor status
```{r}
is.data.frame(dados)
is.factor(dados$trat)
is.numeric(dados$prod)
```

Load necessary package for plotmeans
```{r}
library(gplots)
```

Plot means with confidence intervals
```{r}
plotmeans(prod ~ trat, data = dados, mean.labels = TRUE, ylim = c(30, 70),
          digits = 3, col = 'blue', connect = FALSE, ci.label = TRUE,
          ylab = 'Production', xlab = 'Treatments', cex = 0.9,
          main = "Means with Confidence Intervals")
```

### 3. Fitting ANOVA model

Since this is a Completely Randomized Design (CRD), the product is explained only by the treatment effects.

```{r}
fit1 <- aov(prod ~ trat, data = dados)
summary(fit1)
```

### 4. Contrasts
```{r}
library(gmodels)
```

**a)Define contrasts**

C1 = mA - mD
```{r}
fit.contrast(fit1, trat, c(1, 0, 0, -1))
```

C2 = mB - mC
```{r}
fit.contrast(fit1, trat, c(0, 1, -1, 0))
```

C3 = mA - mB - mC + mD
```{r}
fit.contrast(fit1, trat, c(1, -1, -1, 1))
```

**(order matters!)**
C4 = mA + mD - mB - mC 
```{r}
fit.contrast(fit1, trat, c(1, -1, -1, 1))
```

Groups of contrasts
```{r}
fit.contrast(fit1, trat,
             rbind('(A,D) vs (B,C)' = c(1, -1, -1, 1),
                   'A vs D' = c(1, 0, 0, -1),
                   'B vs C' = c(0, 1, -1, 0)))
```

Contrast matrix
```{r}
cmat <- rbind('(A,D) vs (B,C)' = c(1, -1, -1, 1),
              'A vs D' = c(1, 0, 0, -1),
              'B vs C' = c(0, 1, -1, 0))

fit.contrast(fit1, trat, cmat)
```
