---
title: "Multiple Comparison"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---

Loading data
```{r}
data <- read.table("DataMilho.txt",head=TRUE)
knitr::kable(data)
```


```{r}
data$trat<-factor(data$trat)

n <- length(data$prod)
```

Residue calculation
```{r}
fit1 <- aov(data$prod ~ data$trat)
anova(fit1)

fit1$fitted.values
residuos <- data.frame(fit1$residuals)
res <- residuos[,1]
```
Residual estimatives
```{r}
xb <- mean(res) 
sx <- sd(res)
```

Normality test
```{r warning=FALSE}
library(car)

qqPlot(res, 
       main = "Q-Q Plot with Confidence Envelope",
       col = "blue",
       pch = 20,
       envelope = 0.95)  # 95% confidence envelope
```

```{r warning=FALSE}
library(nortest)
t1 <- ks.test(res, "pnorm", xb, sx)
t2 <- lillie.test(res)             
t3 <- cvm.test(res)                
t4 <- shapiro.test(res)           
t5 <- sf.test(res)                
t6 <- ad.test(res)                
```

Table of results
```{r}
tests <- c(t1$method, t2$method, t3$method, 
            t4$method, t5$method, t6$method)

st <- as.numeric(c(t1$statistic, t2$statistic, t3$statistic,
                     t4$statistic, t5$statistic, t6$statistic))

pv <- c(t1$p.value, t2$p.value, t3$p.value, 
            t4$p.value, t5$p.value, t6$p.value)

results <- data.frame(Test = tests, Statistics = st, pvalue = pv)
knitr::kable(results)
```

Homoscedasticity test
```{r}
boxplot(res ~ data$trat, 
        col="gray", ylab="Residuals", xlab="Varieties")

mean.res <-tapply(res, data$trat, mean)         # Mean for each treat
sd.res <- tapply(res, data$trat, sd)

points(mean.res, pch=20, col=1, cex=1.5)
points(mean.res+sd.res, pch="*", cex=1.5)
points(mean.res-sd.res, pch="*", cex=1.5)
```
With the function `points()` we can set the average over the boxplot

```{r}
bartlett.test(res, data$trat)
```

Checking the independence

```{r}
plot(res,
     type = "b",
     pch = 20,
     col = "blue",
     xlab = "Observation Order",
     ylab = "Residuals",
     main = "Residuals in Order of Collection")
abline(h = 0, col = "red", lty = 2)
```

### 1. Multiple comparisons

```{r}
fit2 <- aov(data$prod~data$trat)
anova(fit2)
```
Once we have run the ANOVA and found a significant effect, we can proceed with multiple comparison tests.

The following tests can be applied after a significant ANOVA result:

-  **Tukey test**
-  **Duncan test** 
-  **SNK test** 
-  **LSD test**
-  **LSDB test** 
-  **SK**

```{r warning=FALSE}
require(ExpDes.pt)
fit3<- dic(data$trat, data$prod, quali = TRUE, 
           mcomp = "tukey", sigT = 0.05, sigF = 0.05)
```

### 2. Orthogonal Contrasts

```{r warning=FALSE}
require(gmodels)
```

Some multiple comparisons can be performed using orthogonal contrasts. Below, we present three examples of such contrasts, which allow us to test specific and independent hypotheses about the treatment means.

C1 = 3mA-mB-mC-mD
```{r}
fit.contrast(fit2, data$trat, c(3, -1, -1,  -1))
```

C2 = mB+mC-2mD
```{r}
fit.contrast(fit2, data$trat, c(0, 1, 1,  -2))
```

C3 = mB-mC
```{r}
fit.contrast(fit2, data$trat, c(0, 1, -1,  0))
```

### 3. Scheffe test
```{r warning=FALSE}
require(DescTools)
```
Scheffe's method applies to the set of estimates of all possible contrasts among the factor level means, not just the pairwise differences considered by Tukey's method.

```{r}
sch2<-ScheffeTest(fit2, contrasts= matrix(c(3,-1,-1,-1,
                                             0, 1, 1,-2,
                                             0, 1,-1,0), 
                                          ncol=3));

plot(sch2)
```
