---
title: "Multiple Comparison"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
ls()
remove(list=ls(all=TRUE))
ls()
```

Packages required

```{r}
install.packages("nortest")
install.packages("outliers")
install.packages("ExpDes.pt")
install.packages("agricolae")
install.packages("SuppDists")
install.packages(c("multcomp","DescTools"))
install.packages("gmodels")
install.packages("laercio")
```

Loading data

```{r}
dados=read.table("DataMilho.txt",head=TRUE)
str(dados)
```


```{r}
dados$trat<-factor(dados$trat)
str(dados)
dados
attach(dados)

n <- length(dados$prod)
cat("\n n =", n, "\n")
```

Residuo calculation

```{r}
fit1 <- aov(prod~trat)
anova(fit1)
names(fit1)
fit1$fitted.values
residuos <- data.frame(fit1$residuals)
res<-residuos[,1]; res
```

Chart res analysis

```{r}
par(mfrow=c(2,2))
plot(fit1)
layout(1)
```

Residual estimatives

```{r}
xb <- mean(res) 
sx <- sd(res)
cat("\n Média amostral =", xb, "\n Desvio padrão amostral =", sx, "\n")
```

Error normality - QQ chart

```{r}
qqn <- qqnorm(res,col="blue", cex=1, main = "", xlab = "Quantis teóricos N(0,1)", pch = 20,
ylab = "Quantis amostrais")
qqline(res, lty = 2, col = "red", lwd=2)
```

QQ chart with envelope

```{r}
nsim <- 500  # Number of simulations
conf <- 0.95 # Reability coenf.

dadossim <- matrix(rnorm(n * nsim, mean = xb, sd = sx), nrow = n)
dadossim <- apply(dadossim, 2, sort)

infsup <- apply(dadossim, 1, quantile, probs = c((1 - conf) / 2,
(1 + conf) / 2))
xbsim <- rowMeans(dadossim); xbsim
```

```{r}
faixay <- range(res, dadossim)
qq0  <- qqnorm(res,col="blue", cex=1, main = "", xlab = "Quantis teóricos N(0,1)", pch = 20,
ylab = "Quantis dos Valores da Amostra", ylim = faixay)
eixox <- sort(qq0$x)
lines(eixox, xbsim,lty = 2, col = "red",lwd=2)
lines(eixox, infsup[1,])
lines(eixox, infsup[2,])
```

Or then

```{r}
require(fBasics)
qqnormPlot(res)
```

Normality test

```{r}
library(nortest)
t1 <- ks.test(res, "pnorm", xb, sx);t1 # Kolmogorov Smirnov
t2 <- lillie.test(res);t2              # Lilliefors
t3 <- cvm.test(res);t3                 # Cramér-von Mises
t4 <- shapiro.test(res);t4             # Shapiro-Wilk
t5 <- sf.test(res);t5                  # Shapiro-Francia
t6 <- ad.test(res);t6                  # Anderson-Darling
```

Table of results

```{r}
testes <- c(t1$method, t2$method, t3$method, t4$method, t5$method,
t6$method)
estt <- as.numeric(c(t1$statistic, t2$statistic, t3$statistic,
t4$statistic, t5$statistic, t6$statistic))
valorp <- c(t1$p.value, t2$p.value, t3$p.value, t4$p.value, t5$p.value,
t6$p.value)
resultados <- cbind(estt, valorp)
rownames(resultados) <- testes
colnames(resultados) <- c("Estatística", "p-value")
print(resultados, digits = 4)
```

Checking the homoscedasticity of the errors.

```{r}
plot(fitted(fit1),pch=19, col=4, resid(fit1),xlab="Valores ajustados (médias trat)", ylab="Resíduos")
```

residuals vs treatment

```{r}
require(lattice)
dotplot(res~trat, pch=19, col=2,ylab="Resíduos",xlab="Tratamentos")
```

Box-plot residuals vs treatment

```{r}
boxplot(res~trat, col=c(2,3,4,5), ylab="Resíduos", xlab="Tratamentos" )
```

With the function "points()" we can set the average over the boxplot

```{r}
mean.res <-tapply(res, trat, mean)         # Mean for each treat
points(mean.res, pch=20, col=1, cex=1.5)
```

In the same way, we can set the standard deviation

```{r}
sd.res <-tapply(res, trat, sd)
points(mean.res+sd.res, pch="*", col=3, cex=2.5)
points(mean.res-sd.res, pch="*", col=3, cex=2.5)
```

```{r}
#4- Gráfico da variavel resposta vs tratamentos (SOMENTE NO DIC)
dotplot(prod~trat, pch=19, col=2,ylab="Produtividade",xlab="Tratamentos")
```

Checking the homoscedasticity of the erros

### F de hartley

Only use when rep is equal for all treats

```{r}
require(SuppDists)
gl = 4
rep = 5
var.res<- tapply(res ,trat,var); var.res
Fmax<-max(var.res)/min(var.res); Fmax
pvalue<-pmaxFratio(Fmax, gl, rep-1, lower.tail=FALSE, log.p=FALSE)
pvalue

detach(package:SuppDists, unload=TRUE) # Descareegar o pacote da memoria
```

Cochran test

```{r}
require(outliers)
d<-data.frame(res, trat);d
h2 <- cochran.test(res~trat, d, inlying=F)
```

Bartlett test

can be used for unbalanced data

```{r}
h3 <- bartlett.test(res, trat)
h3
```

results table

```{r}
homo = c(F-máximo de Hartley, h2$method, h3$method); homo
estt2 = as.numeric(c(Fmax, h2$statistic, h3$statistic)); estt2
pvalor2 = c(pvalue, h2$p.value, h3$p.value); pvalor2
awns = cbind(estt2, pvalor2)
rownames(awns) = homo 
colnames(awns) = c("Estatística", "p-value")
print(awns, digits = 4)
```

Checking the independence of the erros

```{r}
plot(fitted(fit1), resid(fit1),pch=19, col=4,xlab="Valores ajustados (médias trat)", ylab="Resíduos")

da<-data.frame(coleta,res);da             
da1<-da[order(coleta, decreasing=F),];da1 

plot(da1$coleta, da1$res,pch=19,type = "o", col=2, xaxt="n",yaxt="n",
 xlab="Ordem de coleta dos dados", ylab="Resíduos")
axis(1, at = seq(0, 20, by = 1))
axis(2, at = seq(-4, 4, by = 1))
abline(h=0, col="blue" )
#Incluir linhas de grade
summary(res)
abline(h=seq(-4,4,0.5),v=seq(1,20,1),lty=3,col="gray")
```

## Multiple comparisons

```{r}
fit2 <- aov(prod~trat)
anova(fit2)
```

Utilized for calculated CV

```{r}
require(agricolae) 
cv.model(fit2) 
```

or then

mcomp can be: tukey, duncan, snk, lsd,lsdb, sk

```{r}
require(ExpDes.pt)
fit3<- dic(trat, prod, quali = TRUE, mcomp = "tukey", sigT = 0.05, sigF = 0.05)
```

```{r}
require(agricolae)
```

Tukey test

```{r}
HSD.test(fit2, "trat",alpha=0.05, group=TRUE, console=TRUE)
HSD.test(fit2, "trat",group=FALSE, console=TRUE)
```

Duncan test

```{r}
duncan.test(fit2, "trat",alpha=0.05, group=TRUE, console=TRUE)
duncan.test(fit2, "trat",group=FALSE, console=TRUE)
```

## Teste t para grupos de contrastes Ortogonais

### C1= 3mA-mB-mC-mD

### C2= mB+mC-2mD

### C3= mB-mC

```{r}
require(gmodels)
```

C1

```{r}
fit.contrast(fit2,trat, c( 3, -1, -1,  -1))
```

C2

```{r}
fit.contrast(fit2,trat, c( 0, 1, 1,  -2))
```

C3

```{r}
fit.contrast(fit2,trat, c( 0, 1, -1,  0))
```

Schffe test

```{r}
require(DescTools)
```

```{r}
sch2<-ScheffeTest(fit2, contrasts= matrix( c(3,-1,-1,-1,
                                             0, 1, 1,-2,
                                             0, 1,-1,0), ncol=3));sch2
plot(sch2)
```
