---
title: "Split-Plot experiment in completely randomized Design"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction - Non sig

This analysis is based on a split-plot experiment conducted under a Completely Randomized Design (CRD).  
The objective was to evaluate the *brix* (sugar content) of mangoes according to variety and the position of the fruits relative to the cardinal points (North, South, East, West).

The researcher collected 4 fruits (one from each cardinal direction) from each of the 3 replicates for each of the 5 mango varieties.  
Using a 5% significance level, we will perform:

- An Analysis of Variance (ANOVA)  
- Duncan's multiple range test (when applicable)

# Cleaning the Environment

```{r}
rm(list = ls())
```

### Loading data
```{r}
data <- read.table("DataManga.txt", header = TRUE)

# Converting relevant variables to factors
data$trat <- as.factor(data$trat)
data$var <- as.factor(data$var)
data$pcard <- as.factor(data$pcard)
data$rep <- as.factor(data$rep)

# Checking the structure of the dataset
str(data)

# Attaching the dataset to the R search path
attach(data)

# Display the dataset
data

# Display the number of observations
n <- length(data$brix)
cat("\n n =", n, "\n")
```

### Assumption checks
Before performing the full analysis, we check the assumptions of
normality and homogeneity of variances
```{r}
# Basic model to evaluate residuals
fit1 <- aov(brix ~ trat)

# ANOVA table
anova(fit1)

# Shapiro-Wilk test for normality of residuals
shapiro.test(residuals(fit1))

# Bartlett's test for homogeneity of variances
bartlett.test(residuals(fit1) ~ trat)
```

### Split-Plot anova
In this split-plot design:

-The main plot factor is the variety
-The subplot factor is the cardinal position
We use the ExpDes.pt package for analysis.
```{r}
# Load the package
require(ExpDes.pt)

# Split-plot analysis in CRD
fit2 <- psub2.dic(var, pcard, rep, brix,
                  quali = c(TRUE, TRUE),
                  mcomp = "duncan",
                  fac.names = c("Variety", "Cardinal Point"),
                  sigT = 0.05, sigF = 0.05)

```

### Alternativa: Manual ANOVA with aov()
```{r}
# ANOVA model with interaction and proper error structure
fit3 <- aov(brix ~ var * pcard + Error(rep:var))

# Display summary
summary(fit3)
```

## Example 2 - Sig
```{r}
remove(list = ls())
```

### Introduction

In this experiment using a split-plot scheme within a **Randomized Complete Block Design (RCBD)** with 4 blocks, the goal was to compare:

- **4 Oat Varieties**: A1, A2, A3, A4  
- **4 Seed Treatments**: B1, B2, B3 (chemical products), and B4 (untreated control)  

The response variable is **oat production**.

### Loading data
```{r}
data <- read.table("DataTratSementes.txt", header = TRUE)

# Converting variables to factors
data$trat <- as.factor(data$trat)
data$var <- as.factor(data$var)
data$sem <- as.factor(data$sem)
data$bloco <- as.factor(data$bloco)

# Checking structure of the dataset
str(data)

# Attach data to the search path
attach(data)

# Show the dataset
data

# Number of observations
n <- length(data$prod)
cat("\n n =", n, "\n")
```

### Assumptions checks
We perform initial ANOVA on the treatments and blocks to extract residual and test for normality
```{r}
# Initial ANOVA model
fit1 <- aov(prod ~ trat + bloco)
anova(fit1)

# Residuals
res <- residuals(fit1)
res

# Residual statistics
xb <- mean(res)
sx <- sd(res)
cat("\n Sample mean =", xb, "\n Sample standard deviation =", sx, "\n")

# Kolmogorov-Smirnov normality test
ks.test(res, "pnorm", xb, sx)
```

### Split-Plot ANOVA in RCBD
In this split-plot design:

-Main plot factor: Variety (var)
-Subplot factor: Seed Treatment (sem)
We use the ExpDes.pt package for performing the correct statistical analysis.

```{r}
# Load required package
require(ExpDes.pt)

# Split-plot analysis using RCBD
fit2 <- psub2.dbc(var, sem, bloco, prod,
                  quali = c(TRUE, TRUE),
                  mcomp = "tukey",
                  fac.names = c("Varieties", "Seed Treatments"),
                  sigT = 0.05, sigF = 0.05)
```

