---
title: "Split-Plot experiment"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---

# Split-plot experiment

## Example 1: Non-significant Interaction (CRD)

This analysis is based on a split-plot experiment conducted under a Completely Randomized Design (CRD). 

The objective was to evaluate the *brix* (sugar content) of mangoes according to variety and the position of the fruits relative to the cardinal points (North, South, East, West).

The researcher collected 4 fruits (one from each cardinal direction) from each of the 3 replicates for each of the 5 mango varieties.  

```{r}
rm(list = ls())

data <- read.table("DataManga.txt", header = TRUE)

data$trat <- as.factor(data$trat)
data$var <- as.factor(data$var)
data$pcard <- as.factor(data$pcard)
data$rep <- as.factor(data$rep)

n <- length(data$brix)

knitr::kable(data)
```

### 1.1 Assumption checks
```{r}
fit1 <- aov(data$brix ~ data$trat)
anova(fit1)
```

Shapiro-Wilk test for normality of residuals
```{r}
shapiro.test(residuals(fit1))
```

Bartlett's test for homogeneity of variances
```{r}
bartlett.test(residuals(fit1) ~ data$trat)
```

### 1.2 Analysis of variance

In this split-plot design:

-   The main plot factor is the variety
-   The subplot factor is the cardinal position

We use the `ExpDes.pt` package for analysis.

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
require(ExpDes.pt)

fit2 <- psub2.dic(data$var, data$pcard, 
                  data$rep, data$brix,
                  quali = c(TRUE, TRUE),
                  mcomp = "duncan",
                  fac.names = c("Variety", 
                                "Cardinal Point"),
                  sigT = 0.05, sigF = 0.05)
```

### 1.3 Alternativa: Manual ANOVA with aov()
```{r}
fit3 <- aov(data$brix ~ data$var * data$pcard + Error(data$rep:data$var))
summary(fit3)
```

## Example 2: Significant Interaction (RCBD)

In this experiment using a split-plot scheme within a **Randomized Complete Block Design (RCBD)** with 4 blocks, the goal was to compare:

-   **4 Oat Varieties**: A1, A2, A3, A4  
-   **4 Seed Treatments**: B1, B2, B3, and B4 (untreated control)  

The response variable is **oat production**.

```{r message=FALSE, warning=FALSE}
remove(list = ls())
data <- read.table("DataTratSementes.txt", header = TRUE)

data$trat <- as.factor(data$trat)
data$var <- as.factor(data$var)
data$sem <- as.factor(data$sem)
data$bloco <- as.factor(data$bloco)

n <- length(data$prod)
knitr::kable(data)
```

### 2.1 Assumptions checks

We perform initial ANOVA on the treatments and blocks to extract residual and test for normality
```{r}
fit1 <- aov(data$prod ~ data$trat + data$bloco)
anova(fit1)
```

Shapiro-Wilk test for normality of residuals
```{r}
shapiro.test(residuals(fit1))
```

Bartlett's test for homogeneity of variances
```{r}
bartlett.test(residuals(fit1) ~ data$trat)
```

### 2.2 Analysis of variance
In this split-plot design:

-Main plot factor: Variety (var)
-Subplot factor: Seed Treatment (sem)
We use the `ExpDes.pt` package for performing the correct statistical analysis.

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
require(ExpDes.pt)

fit2 <- psub2.dbc(data$var, data$sem, data$bloco, data$prod,
                  quali = c(TRUE, TRUE),
                  mcomp = "tukey",
                  fac.names = c("Varieties", 
                                "Seed Treatments"),
                  sigT = 0.05, sigF = 0.05)
```

