---
title: "LinearRegression-SD"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---

# Second-degree Linear Regression
Dry matter (g/plant) as a function of zinc doses (ppm)
```{r}
ls()           # List all created objects
rm(list=ls())  # Remove all created objects
ls()

x <- c(1.0, 2.5, 4.0, 5.5, 7.0, 8.5)
y <- c(20.3, 31.3, 34.6, 35.1, 30.2, 19.7)

ms <- data.frame(x, y)
ms
attach(ms)

plot(y ~ x, col = 4, pch = 19, data = ms,
     main = "Scatter plot: Dry matter vs Zinc dose",
     xlab = "Zinc dose (ppm)", ylab = "Dry matter (g/plant)")

```

## Model selection and assumption checking

Fit first-degree model

```{r}
fit1 <- lm(y ~ x)
summary(fit1)
anova(fit1)
```

The first-degree model does not adequately explain the variability in y.

Fit second-degrre model

```{r}
fit2 <- update(fit1, . ~ . + I(x^2))  # Option 1
summary(fit2)
anova(fit2)

fit3 <- lm(y ~ x + I(x^2))  # Option 2
summary(fit3)
anova_fit3 <- anova(fit3)
print(anova_fit3, digits = 6)

fit4 <- lm(y ~ poly(x, degree = 2, raw = TRUE))  # Option 3
summary(fit4)
anova_fit4 <- anova(fit4)
print(anova_fit4, digits = 6)
```

Assumption: normality
```{r}
shapiro.test(fit3$res)
```

Adjust regresstion plot
```{r}
plot(y ~ x, data = ms, pch = 16, col = "blue", xaxt = "n", 
     xlab = "Zinc dose (ppm)", ylab = "Dry matter (g/plant)", ylim = c(18, 40))
axis(1, xaxp = c(0, 10, 20))

cf.m4 <- coef(fit3)
curve(cf.m4[1] + cf.m4[2]*x + cf.m4[3]*x^2, add = TRUE, col = "red", lwd = 2)

co <- format(c(coef(fit3), R2), digits = 4)
text(4.7, 27, label = substitute(hat(y[i])==b0+b1*x[i]+b2*x[i]^2~~~~R^2==r2,
                                 list(b0 = co[1], b1 = co[2], b2 = co[3], r2 = co[4])),
     bty = "n", cex = 0.8)
```

Vertex and ferivate analysis
```{r}
# First derivative: find critical point
D(expression(11.24222 + 10.4677*x - 1.113492*x^2), "x")

# Solve for the root (critical point)
ro <- uniroot(function(x) 10.4677 - 1.113492 * (2 * x), c(1, 8.5))
str(ro)

abline(v = ro$root, h = 0, col = 1, lty = 3)

# Maximum dry matter value
xc <- ro$root
xc
y_max <- 11.24222 + 10.4677*xc - 1.113492*xc^2
y_max
abline(v = xc, h = y_max, col = 1, lty = 3)
```

Second derivative test
```{r}
# Second derivative to confirm maximum point
D(D(expression(11.24222 + 10.4677*x - 1.113492*x^2), "x"), "x")
```
Since the second derivative is negative, the critical point is a maximum.
