---
title: "Linear Regression with Repeated Measures in CRD (Completely Randomized Design)"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
# Clean environment
rm(list = ls())
ls()

# Load data
data <- read.table("3EST220-DataSonifero.txt", header = TRUE)
head(data)
attach(data)
```

Assumptions checking (CRD)
```{r}
# Linear model treating x as a qualitative variable
fit1 <- lm(y ~ factor(x))
anova(fit1)

# Residuals
res <- fit1$residuals
res

# Normality test
shapiro.test(res)

# Homogeneity of variances
bartlett.test(res, x)
```

Regression model (1st-degree)
```{r}
# Scatterplot
plot(x, y, pch = 19, col = "red", ylim = c(0, 20), main = "Scatterplot")
plot(tapply(y, x, mean) ~ unique(x), pch = 19, col = "red", ylim = c(0, 20),
     main = "Mean Response per Dose")

# Model fitting
fit2 <- lm(y ~ x)
summary(fit2)
R2 <- summary(fit2)$r.squared; R2
anova(fit2)
coef(fit2)
confint(fit2)
```

Lack of fit and pure error
```{r}
# Fit model treating x as a factor
fit3 <- lm(y ~ factor(x))
summary(fit3)
R2max <- summary(fit3)$r.squared; R2max
anova(fit3)

# Lack of fit test
anova(fit2, fit3)
```

Interpretation:

Total residual sum of squares (RSS) = 112
Pure error = 100
Lack of fit = 12 (df = 3, F = 0.4, p = 0.7561)
Not significant → model fits well

```{r}
# Real R-squared
r2real <- R2 / R2max
r2real
```

Using expdes.pt package
```{r}
library(ExpDes.pt)

# Quantitative data (quali = FALSE)
dic(x, y, quali = FALSE, mcomp = "tukey", sigT = 0.05, sigF = 0.06)
```

### Polynomial regression via reg.poly
```{r}
# Get ANOVA info
dic <- lm(y ~ factor(x))
anova(dic)

# Apply polynomial regression
reg.poly(y, x, 10, 100, 4, 132)
```

### Regression plot with equation
```{r}
# Plot data points
plot(y ~ x, data = data, pch = 16, col = "blue",
     xlab = "Dose of sedative (mg)", ylab = "Sleep duration (hours)",
     ylim = c(0, 20), frame = FALSE)

# Add regression line
cf.m <- coef(fit2)
curve(cf.m[1] + cf.m[2]*x, add = TRUE, col = "red", lty = 1)

# Annotate with equation and R²
co <- format(c(coef(fit2), r2real), digits = 4)
text(3.5, 5.0,
     label = substitute(hat(y[i]) == b0 + b1*x[i] ~~~~ r^2 == r2,
                        list(b0 = co[1], b1 = co[2], r2 = co[3])))
box()
```

### Fitting with mean and totals
```{r}
# Using mean values
y1 <- c(5, 9, 10, 13, 13)
x1 <- c(1, 2, 3, 4, 5)
fit5 <- lm(y1 ~ x1)
summary(fit5)
anova(fit5)

# Using total values
y2 <- c(15, 27, 30, 39, 39)
x2 <- c(1, 2, 3, 4, 5)
fit6 <- lm(y2 ~ x2)
summary(fit6)
anova(fit6)
```

Note:

Both models yield the same regression equation form
When using totals, coefficients are multiplied by the number of replicates
Sums of squares also differ by a factor equal to the number of replicates
