---
title: "Hypothesis Tests"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---

### 1. Clean up the environment
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
rm(list=ls())  # Remove any previously created objects
ls()           # Check the objects in the current R session
```

### 2. Installing and loadind packages
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# install.packages(c("ggplot2", "dplyr"))

library(ggplot2)
library(dplyr)
```

### 3. Save and load results

Save created objects
```{r}
save(list = ls(all.names = TRUE), file = "Result.RData", envir = .GlobalEnv)
```

Load saved file after exiting R
```{r}
load("Result.RData", envir = parent.frame(), verbose = FALSE)
```

Check objects in the "Result.RData" file
```{r}
ls()
```

## Example 1: F-test for Variance

The quality of bricks is considered better when their homogeneity is higher. Six bricks from two brands were tested for shear strength, and rupture forces were measured. The producer of Brand B claims their bricks are better. Use a 5% significance level.

```{r}
brand_A <- c(34.9, 35.5, 38.8, 39.2, 33.7, 37.6)
brand_B <- c(38.5, 39.0, 40.7, 42.9, 37.8, 41.4)

var(brand_A)
var(brand_B)
```

**a) F-test for variances**

The null hypothesis is that the ratio of variances is equal to 1, and the alternative is that it is not equal to 1.
```{r}
var.test(brand_A, brand_B, ratio = 1, alternative = "greater", conf.level = 0.95)
```
Another way to calculate the p-value

```{r}
pf(1.326, df1 = 5, df2 = 5, lower = FALSE, log.p = FALSE)
```

## Example 2: t-Test for single mean

In healthy individuals, renal oxygen consumption is normally distributed around 12 cm³/min.We want to test if a certain disease affects this value using data from five patients. Test at a 1% significance level.

```{r}
o2 = data.frame(v1 = as.factor(1),
                v2 = c(14.4, 12.9, 15.0, 13.7, 13.5))
```

Using the `t.test()` function, we can perform a t-test for a mean when the argument paired is set to `FALSE`.

**a) Testing for single mean**
```{r}
t = t.test(o2$v2, 
           mu = 12, 
           paired = FALSE, 
           alternative = "two.sided", 
           conf.level = 0.99);t
```
Another way to calculate the p-value is by using the `pt()` function, providing the test statistic, degrees of freedom, and multiplying by 2, since it is a two-tailed test.

```{r}
pt(5.2099, df = 4, lower = FALSE) * 2
```
We can also access the values by referring to the object `t` created when using `t.test()`.

When manipulating data, another way to access specific values, besides using the `$` operator, is by using double square brackets `[[]]`.

```{r}
pt(t[[1]], df = t[[2]], lower = F) * 2
```
Repeat the test with a 95% confidence interval

```{r}
teste_t <- t.test(o2$v2, 
                  mu = 12,
                  paired = FALSE, 
                  alternative = "two.sided", 
                  conf.level = 0.95);teste_t
```

**b) Boxplot for the sample**

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
mean_v2 <- mean(o2$v2)
sd_v2 <- sd(o2$v2)

ggplot(o2, aes(x = v1, y = v2)) +
  geom_boxplot() +
  geom_point(aes(y = mean(v2)), color = "red", size = 3) +
  geom_text(aes(y = mean(v2), label = "Mean (p-value = 0.0065)"),
            color = "black", size = 3, vjust = 2) +
  geom_point(aes(y = mean(v2) + sd(v2)), color = "blue", size = 3) +
  geom_point(aes(y = mean(v2) - sd(v2)), color = "blue", size = 3) +
  labs(y = "Oxygen consumption")
```
## Example 3: t-Test for Two Independent Means (Case A)

Data represents the strength of two types of concrete. At a 5% significance level, we want to test if concrete 1 is stronger than concrete 2.

```{r}
c1 <- c(54, 55, 58, 51, 57)
c2 <- c(50, 54, 56, 52, 53)
```

**a) Test for variance homogeneity**
```{r}
var(c1)
var(c2)

var.test(c1, c2, ratio = 1, alternative = "greater", conf.level = 0.95)
```
**b) Conduct the t-test**

```{r}
t <- t.test(c1, c2, 
       mu = 0, 
       paired = FALSE, 
       var.equal = TRUE, 
       alternative = "greater", 
       conf.level = 0.95)
```

**c) Plotting two samples test**

With R's `boxplot()` function, we can easily create boxplot charts.
```{r}
df <- data.frame(name = as.factor(rep(1:2, each = 5)),
                values = c(c1,c2))

boxplot(values ~ name,
        data = df,
        border = "black",
        ylab = "values",
        xlab = "Concretes")

means <- tapply(df$values, df$name, mean)
sds   <- tapply(df$values, df$name, sd)

points(1:2, means, pch = 19, col = "red", cex = 1.5)
points(1:2, means + sds, pch = 19, col = "blue", cex = 1.5)
points(1:2, means - sds, pch = 19, col = "blue", cex = 1.5)

text(x = 1, y = 55.5, labels = "55", col = "black")
text(x = 2, y = 53.5, labels = "53", col = "black")
text(x = 1.5, y = 58, labels = paste0("p-valor = ", round(t$p.value, 2)))
```

## Example 4: t-Test for Two Independent Means (Case B)

Comparing salaries of two professional groups: general liberal professionals and business administration graduates.

Test at a 5% significance level.

```{r}
lib <- c(6.6, 10.3, 10.8, 12.9, 9.2, 12.3, 7.0)
adm <- c(8.1, 9.8, 8.7, 10.0, 10.2, 10.8, 8.2, 8.7, 10.1)
```

**a) Testing for variance homogeneity**
```{r}
var(lib)
var(adm)

var.test(lib, 
         adm, 
         ratio = 1, 
         alternative = "greater", 
         conf.level = 0.95)
```
**b) Conduct the t-test**
```{r}
TST <- t.test(lib, adm, 
              mu = 0, 
              paired = FALSE, 
              var.equal = FALSE, 
              alternative = "two.sided", 
              conf.level = 0.95)
TST
```
**c) Plotting boxplot with mean ± SD**
```{r}
df2 = data.frame(name = factor(rep(c("Lib.", "Adm."), c(7,9))),
                 values = c(lib, adm))

boxplot(values~name,
        data = df2, 
        xlab = "Groups", 
        ylab = "Values", 
        border = "black", 
        outline = TRUE)

means <- tapply(df2$values, df2$name, mean)
sds <- tapply(df2$values, df2$name, sd)

points(1:2, means, pch = 19, col = "red", cex = 1.5)
points(1:2, means + sds, pch = 19, col = "blue", cex = 1.5)
points(1:2, means - sds, pch = 19, col = "blue", cex = 1.5)

text(x = 1, y = 9, labels = "9.4", col = "black")
text(x = 2, y = 9, labels = "9.87", col = "black")
text(x = 1.5, y = 12, labels = paste0("p-valor = ", round(TST$p.value, 2)))

```

## Example 5: t-Test for Paired Data (Before vs After)

We want to test if a chemical product is effective in repelling household insects. The number of insects was counted before and after the application of the product.

```{r}
before <- c(8, 6, 7, 8, 9, 6, 7)
after <- c(4, 0, 3, 5, 3, 4, 2)
```

**a) Testing for paired data**

As we can see from the other performed t-tests, here we set paired equal to `TRUE`.
```{r}
t.test(after, before, paired = TRUE, alternative = "less", conf.level = 0.95)
```

**b) Plotting the Differences**
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
df = data.frame(name = factor(rep(c("before", "after"), c(7, 7))),
                values = c(before, after))

boxplot(values ~name,
        data = df, 
        xlab = "Groups", 
        ylab = "Values", 
        border = "black", 
        outline = TRUE)

means <- tapply(df$values, df$name, mean)
sds <- tapply(df$values, df$name, sd)

points(1:2, means, pch = 19, col = "red", cex = 1.5)
points(1:2, means + sds, pch = 19, col = "blue", cex = 1.5)
points(1:2, means - sds, pch = 19, col = "blue", cex = 1.5)
```
