---
title: "Latin Square Design"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---

# Latin Square Design

A researcher designed an experiment to compare 5 types of bacilli (A, B, C, D, and E) used in yogurt production. However, the researcher found that the 25 available 1-liter milk samples were not homogeneous due to variations in fat content and acidity level. To control these two sources of variation, the bacilli were randomly assigned to the milk samples such that each bacillus was tested under all conditions.

```{r}
rm(list=ls())

data <- read.table("DataIogurte.txt", header=TRUE)

data$trat <- factor(data$trat)
data$linha <- factor(data$linha)
data$coluna <- factor(data$coluna)

n <- length(data$volume)

knitr::kable(data)
```

### 1. ANOVA and residuals
```{r}
fit1 <- aov(data$volume ~ data$linha + data$coluna + data$trat)
print(anova(fit1), digits=7)
res <- residuals(fit1)
```

### 2. Means and variances
```{r}
model.tables(fit1, "means")
var.trat <- tapply(data$volume, data$trat, var)
var.res <- tapply(res, data$trat, var)
```

### 3. Preposition residual verification
```{r}
xb <- mean(res)
sx <- sd(res)
```

**3.1. Checking residuals normality distribution**
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(car)

qqPlot(res, 
       main = "Q-Q Plot with Confidence Envelope",
       col = "blue",
       pch = 20,
       envelope = 0.95)  # 95% confidence envelope
```
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(nortest)
tests <- list(ks = ks.test(res, "pnorm", xb, sx),
              lillie = lillie.test(res),
              cvm = cvm.test(res),
              shapiro = shapiro.test(res),
              sf = sf.test(res),
              ad = ad.test(res))

results <- data.frame(Statistic = sapply(tests, function(x) as.numeric(x$statistic)),
                      P_Value = sapply(tests, function(x) x$p.value),
                      row.names = sapply(tests, function(x) x$method))

knitr::kable(results)
```

**3.2 Checking the residuals independence**
```{r}
plot(res,
     type = "b",
     pch = 20,
     col = "blue",
     xlab = "Observation Order",
     ylab = "Residuals",
     main = "Residuals in Order of Collection")
abline(h = 0, col = "red", lty = 2)
```

**3.3 Checking residuals homoscedasticity**
```{r}
bartlett.test(res, data$trat)
```

### 4. ANOVA for Latin Square
```{r}
fit2 <- aov(data$volume ~ data$linha + data$coluna + data$trat)
print(anova(fit2), digits=7)
```

Using ExpDes.pt
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(ExpDes.pt)
fit3 <- dql(data$trat, data$linha, data$coluna, data$volume, 
            quali=TRUE, mcomp="tukey", sigT=0.05, sigF=0.05)
```
