---
title: "Latin Square Design"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---

# Latin Square Design
A researcher designed an experiment to compare 5 types of bacilli
(A, B, C, D, and E) used in yogurt production. However, the researcher
found that the 25 available 1-liter milk samples were not homogeneous
due to variations in fat content and acidity level. To control these two
sources of variation, the bacilli were randomly assigned to the milk
samples such that each bacillus was tested under all conditions.

### Clearing all objects from memory
```{r}
rm(list=ls())
ls()
```

### Loading data
```{r}
data <- read.table("DataIogurte.txt", header=TRUE)
str(data)
data$trat <- factor(data$trat)
data$linha <- factor(data$linha)
data$coluna <- factor(data$coluna)
str(data)
data
attach(data)

n <- length(volume)
cat("\n n =", n, "\n")
```

### ANOVA and residuals
```{r}
fit1 <- aov(volume ~ linha + coluna + trat)
print(anova(fit1), digits=7)
res <- residuals(fit1)
```

### Means and variances
```{r}
model.tables(fit1, "means")
var.trat <- tapply(volume, trat, var)
var.res <- tapply(res, trat, var)
```

### Coefficient of Variation (CV)
```{r}
anova_table <- anova(fit1)
CV <- (sqrt(anova_table$Mean[4]) / model.tables(fit1, "means")$tables$Gran) * 100
print(CV, digits=3)
```

### Residual statistics
```{r}
xb <- mean(res)
sx <- sd(res)
cat("\n Residual mean =", xb, "\n Residual SD =", sx, "\n")
```

### Normality verification
```{r}
qqnorm(res, col="blue", cex=1, pch=20, main="", xlab="Theoretical Quantiles N(0,1)", ylab="Sample Quantiles")
qqline(res, lty=2, col="red", lwd=2)
```

### QQ plot with envelope
```{r}
nsim <- 500
conf <- 0.95
data_sim <- matrix(rnorm(n * nsim, mean = xb, sd = sx), nrow = n)
data_sim <- apply(data_sim, 2, sort)
infsup <- apply(data_sim, 1, quantile, probs = c((1 - conf) / 2, (1 + conf) / 2))
xbsim <- rowMeans(data_sim)
y_range <- range(res, data_sim)
qq0 <- qqnorm(res, col="blue", cex=1, main="", pch=20, ylim=y_range, xlab="Theoretical Quantiles N(0,1)", ylab="Sample Quantiles")
x_axis <- sort(qq0$x)
lines(x_axis, xbsim, lty=2, col="red", lwd=2)
lines(x_axis, infsup[1,])
lines(x_axis, infsup[2,])
```

### Normality tests
```{r}
library(nortest)
tests <- list(
  ks = ks.test(res, "pnorm", xb, sx),
  lillie = lillie.test(res),
  cvm = cvm.test(res),
  shapiro = shapiro.test(res),
  sf = sf.test(res),
  ad = ad.test(res)
)

results <- data.frame(
  Statistic = sapply(tests, function(x) as.numeric(x$statistic)),
  P_Value = sapply(tests, function(x) x$p.value),
  row.names = sapply(tests, function(x) x$method)
)
print(results, digits=4)
```

### Homoscedasticity checks
```{r}
plot(fitted(fit1), resid(fit1), pch=19, col=4, xlab="Fitted Values", ylab="Residuals")
plot.default(trat, res, pch=19, col=4, xlab="Treatments", ylab="Residuals")
boxplot(res ~ trat, col=c(2,7,4,5,6), xlab="Treatments", ylab="Residuals")
points(tapply(res, trat, mean), pch=20, col=1, cex=1.5)
sd.res <- tapply(res, trat, sd)
points(tapply(res, trat, mean) + sd.res, pch="-", col=3, cex=2.5)
points(tapply(res, trat, mean) - sd.res, pch="-", col=3, cex=2.5)
```

### Hartley’s F-max test
```{r}
library(SuppDists)
var_res <- tapply(res, trat, var)
Fmax <- max(var_res) / min(var_res)
p_value <- pmaxFratio(Fmax, 4, 5, lower.tail=FALSE, log.p=FALSE)
cat("Fmax =", Fmax, ", p-value =", p_value, "\n")
```

### Cochran’s test
```{r}
library(outliers)
d <- data.frame(res, trat)
cochran.test(res ~ trat, d, inlying=FALSE)
```

### Bartlett’s test
```{r}
bartlett.test(res, trat)
```

### Independence of errors
```{r}
da <- data.frame(coleta, res)
da1 <- da[order(coleta), ]
plot(da1$coleta, da1$res, pch=19, type="o", col=2, xaxt="n", yaxt="n", xlab="Order of Data Collection", ylab="Residuals")
axis(1, at=seq(0, 25, 1))
axis(2, at=seq(-80, 100, 20))
abline(h=0, col="blue")
abline(h=seq(-80,100,20), v=seq(0,25,1), lty=3, col="gray")
```

### ANOVA for Latin Square
```{r}
fit2 <- aov(volume ~ linha + coluna + trat)
print(anova(fit2), digits=7)
```

### Post-hoc tests
```{r}
library(agricolae)
HSD.test(fit2, "trat", group=TRUE, console=TRUE)
duncan.test(fit2, "trat", group=TRUE, console=TRUE)
```

### Using ExpDes.pt
```{r}
library(ExpDes.pt)
fit3 <- dql(trat, linha, coluna, volume, quali=TRUE, mcomp="tukey", sigT=0.05, sigF=0.05)
```
Options for mcomp: "tukey", "duncan", "snk", "lsd", "lsdb", "sk"
