---
title: "Analysis of variance"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
rm(list = ls())
```

# Analysis of variance (ANOVA)

### Exemple 1: Variety of wheat

```{r}
var <- rep(1:2,each=5);var

prod <- c(2.0,2.2,2.3,2.5,3.0,3.2,2.8,2.9,2.4,2.7);prod

dados <- data.frame(var, prod);dados
```

Calculating mean
```{r}
mean.prod <- tapply(prod, var, mean); mean.prod
```

Calculating variance
```{r}
var.prod <-tapply(prod, var, var); var.prod     
```

Calculating standard deviation
```{r}
sd.prod <- tapply(prod, var, sd); sd.prod      
```

The `aov()` function is used for **balanced** data.
```{r}
fit1 <- aov(prod~factor(var), data = dados)
anova(fit1)

summary(fit1)
```

Fitting a linear model using `lm()` function, which can be used for **balanced** and **unbalanced** data.
```{r}
fit2 <- lm(prod~factor(var), data = dados)
anova(fit2)
```

Extracting the predicted values from the model using `fitted.values()`
```{r}
yh <- fitted.values(fit1);yh
```

Getting residual values from the fitted model using `residuals()`
```{r}
res <- residuals(fit1); res
```

Residual standardized
```{r}
res_Stand <- rstandard(fit1); res_Stand
round(data.frame(res,res_Stand),5)

(prod-yh) # other way to get res
```
