---
title: "Completed randomized design"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---

Loading data

```{r}
data <- data.frame(rep = rep(1:5, each=4),
                   trat = factor(rep(c("A", "B", "C", "D"),
                                        times=5)),
                   yield = c(25, 31, 22, 33,
                             26, 25, 26, 29,
                             20, 28, 28, 31,
                             23, 27, 25, 34,
                             21, 24, 29, 28))

knitr::kable(data)
```

### 1. Creating sketch

When we have a daily chore or an activity that is performed frequently, we can create functions to save time and reduce repetitive work.

By utilizing the syntax `function(x) {}`, we can define the behavior we want to automate.

In this example, I created a function to combine all positions between the number of rows and columns using the `expand.grid()` function. Then, to follow one of the principles of experimental design — **randomization** — I used the `sample()` function to randomly assign positions within the created object.

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(dplyr)
library(desplot)

generate_positions <- function(n_cols, n_rows) {
  expand.grid(col = 1:n_cols, row = 1:n_rows) %>%
    sample_frac()
}

generate_positions_by_block <- function(n_rows, n_cols) {
  expand.grid(row = 1:n_rows, col = 1:n_cols)
}

field_layout <- function(data, n_col, n_row, design) {
  data$rep <- as.factor(data$rep)
  data$trat <- as.factor(data$trat)
  sketch <- list()

  if (design == "DIC" || design == "CRD") {
    positions <- generate_positions(n_col, n_row)
    data <- data %>%
      mutate(col = positions$col, row = positions$row)

    plot <- desplot(trat ~ col * row,
                    data = data,
                    text = trat,
                    main = "Field Layout - CRD",
                    show.key = FALSE,
                    cex = 0.8,
                    show.legend = NA,
                    col.regions = colorRampPalette(c("#d7191c", 
                                                     "#fdae61", 
                                                     "#ffffbf", 
                                                     "#a6d96a",
                                                     "#1a9641"))(nlevels(data$trat)))
}

  else if (design == "DBC" || design == "RCBD") {
    colnames(data)[colnames(data) == "bloco"] <- "rep"
    data <- arrange(data, rep)

    n_trat <- nlevels(data$trat)
    n_rep <- nlevels(data$rep)
    n_plots <- n_trat * n_rep

    if ((n_row * n_col) %% n_rep != 0)
      stop("Number of rows and columns must result in equal-sized blocks.")

    plots_per_block <- n_plots / n_rep
    block_cols <- n_col / n_rep
    block_rows <- n_row

    positions_list <- vector("list", n_rep)
    for (i in 1:n_rep) {
      block_pos <- generate_positions_by_block(block_rows, block_cols)
      block_pos$col <- block_pos$col + (i - 1) * block_cols
      positions_list[[i]] <- block_pos[sample(nrow(block_pos)), ]
    }

    all_positions <- do.call(rbind, positions_list)
    data <- data[order(data$rep), ]
    data$col <- all_positions$col
    data$row <- all_positions$row

    plot <- desplot(trat ~ col + row | rep,
                    data = data,
                    text = trat,
                    cex = 0.7,
                    shorten = "no",
                    show.legend =F,
                    out1 = rep,
                    main = "Field Layout - RCBD",
                    col.regions = colorRampPalette(c("#d7191c", 
                                                     "#fdae61", 
                                                     "#ffffbf",
                                                     "#a6d96a", 
                                                     "#1a9641"))(nlevels(data$trat)))
  }
  sketch$plot <- plot
  sketch$data <- data
  return(sketch)
}
```


```{r}
set.seed(10)

field_layout(data = data,
             n_row = 4,
             n_col = 5,
             design = "DIC")
```

### 1. ANOVA Analysis

As we have seen, by using the `aov()` function, we can model the ANOVA for the *yield* through the treatment effects, which correspond to the *trat*.

```{r}
fit <- aov(yield ~ trat, data = data)

anova(fit)
```

### 2. Residual Analysis

Residual analysis is performed because the assumptions of ANOVA are related to the residual values.

Remember the assumptions of ANOVA:

-   **Additivity:** The effects of the different factors in the model are additive, meaning they do not interact in a non-linear way.

-   **Normality of residuals:** The residuals should follow a normal distribution.

-   **Homogeneity of variance:** The variance of residuals should be constant across groups (also known as homoscedasticity).

-   **Independence of observations:** The observations are independent of each other.

```{r}
residuals <- fit$residuals
mean_res <- mean(residuals)
sd_res <- sd(residuals)
```

### 3. Normality Tests

Installing and loading the `"nortest"` package, which is used for a trat of normality tests, such as:

-   Kolmogorov-Smirnov Tests - `ks.test()`
-   Lilliefors Test - `lillie.test()`
-   Cramer-von Mises test - `cvm.test()`
-   Shapiro-Wilk Normality Test - `shapiro.test()`
-   Shapiro-Francia test - `sf.test()`
-   Anderson-Darling test - `ad.test()`

```{r warning=FALSE}
# install.packages(nortest)
library(nortest)

t1 <- ks.test(residuals, "pnorm", mean_res, sd_res)
t2 <- lillie.test(residuals)
t3 <- cvm.test(residuals)
t4 <- shapiro.test(residuals)
t5 <- sf.test(residuals)
t6 <- ad.test(residuals)
```

Taking the name, statistics, and p-values from each performed normality test and organizing them into a dataframe.

```{r}
tests <- c(t1$method, t2$method, t3$method, 
           t4$method, t5$method, t6$method)
statistics <- c(t1$statistic, t2$statistic, t3$statistic, 
                t4$statistic, t5$statistic, t6$statistic)
p_values <- c(t1$p.value, t2$p.value, t3$p.value, 
              t4$p.value, t5$p.value, t6$p.value)
```

```{r}
results <- data.frame(Test = tests, Statistic = statistics, P_Value = p_values)
knitr::kable(results)
```

### 4. Homoscedasticity Tests

```{r fig.height=6, fig.width=10}
boxplot(residuals ~ data$trat, 
        col="gray", ylab="Residuals", xlab="Varieties")
```

Installing and loading the package `lawstat` we can have acess to the function `levene.test()` to perform **Levene's Test of Equality of Variances**

```{r warning=FALSE}
# install.packages("lawstat")
library(lawstat)
levene.test(residuals, data$trat)
```

### 5. Independence Check

It\`s looks difficult to conclude something right?
```{r fig.height=5, fig.width=10}
plot(residuals,
     type = "b",
     pch = 20,
     col = "blue",
     xlab = "Observation Order",
     ylab = "Residuals",
     main = "Residuals in Order of Collection")
abline(h = 0, col = "red", lty = 2)
```

### 6. Multiple Comparisons

By installing and loading the package `ExpDes` we can speed up the process, utilizing `dic()` for and Completed Ramdomized Design (CMD)

```{r warning=FALSE}
# install.packages("ExpDes.pt")
library(ExpDes.pt)

dic(data$trat, data$yield, quali = TRUE, 
    mcomp = "lsdb", hvar= "bartlett", sigT = 0.05, sigF = 0.05)
```

### 7. CRD - Unbalanced Data

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
remove(list=ls())
source("ExperimentalDesignFunction.R")
```

```{r}
data_unbalanced <- read.table("DataUnbalance.txt", head=TRUE)
data_unbalanced$trat <- factor(data_unbalanced$trat)

n <- length(data_unbalanced$result)
```

Plotting field layout
```{r}
set.seed(7)
layout <- field_layout(data = data_unbalanced,
             n_col = 3,
             n_row = 5, 
             design = "DIC")
layout$plot
```
```{r}
knitr::kable(layout$data)
```

```{r}
fit_unbalanced <- aov(result ~ trat, data = data_unbalanced)
anova(fit_unbalanced)
```

```{r}
residuals_unbalanced <- fit_unbalanced$residuals
mean_res_unbalanced <- mean(residuals_unbalanced)
sd_res_unbalanced <- sd(residuals_unbalanced)
```

```{r fig.height=5, fig.width=10, message=FALSE, warning=FALSE, paged.print=FALSE}
# install.packages("car")
library(car)

qqPlot(residuals_unbalanced, 
       main = "Q-Q Plot with Confidence Envelope",
       col = "blue",
       pch = 20,
       envelope = 0.95)  # 95% confidence envelope
```

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(nortest)
t1 <- ks.test(residuals_unbalanced, 
              "pnorm", mean_res_unbalanced, sd_res_unbalanced)
t2 <- lillie.test(residuals_unbalanced)
t3 <- cvm.test(residuals_unbalanced)
t4 <- shapiro.test(residuals_unbalanced)
t5 <- sf.test(residuals_unbalanced)
t6 <- ad.test(residuals_unbalanced)
```

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
tests <- c(t1$method, t2$method, t3$method, t4$method, t5$method, t6$method)
statistics <- c(t1$statistic, 
                t2$statistic, 
                t3$statistic, 
                t4$statistic, 
                t5$statistic, 
                t6$statistic)

p_values <- c(t1$p.value, t2$p.value, t3$p.value, 
              t4$p.value, t5$p.value, t6$p.value)
```

```{r}
results <- data.frame(Test = tests, 
                      Statistic = statistics, 
                      P_Value = p_values)

knitr::kable(results)
```

```{r}
fit_final <- aov(result ~ trat, data = data_unbalanced)
anova(fit_final)
```

```{r message=FALSE, warning=FALSE}
library(ExpDes.pt)
dic(data_unbalanced$trat, data_unbalanced$result, quali = TRUE, mcomp = "tukey", sigT = 0.05, sigF = 0.05)
```
