---
title: "Completed randomized design"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---

Loading data

```{r}
data <- data.frame(rep = rep(1:5, each=4),
                   variety = factor(rep(c("A", "B", "C", "D"), times=5)),
                   yield = c(25, 31, 22, 33,
                             26, 25, 26, 29,
                             20, 28, 28, 31,
                             23, 27, 25, 34,
                             21, 24, 29, 28))

knitr::kable(data)
```

### 1. ANOVA Analysis

As we have seen, by using the `aov()` function, we can model the ANOVA for the *yield* through the treatment effects, which correspond to the *variety*.

```{r}
fit <- aov(yield ~ variety, data = data)

anova(fit)
```

### 2. Residual Analysis

Residual analysis is performed because the assumptions of ANOVA are related to the residual values.

Remember the assumptions of ANOVA:

-   **Additivity:** The effects of the different factors in the model are additive, meaning they do not interact in a non-linear way.

-   **Normality of residuals:** The residuals should follow a normal distribution.

-   **Homogeneity of variance:** The variance of residuals should be constant across groups (also known as homoscedasticity).

-   **Independence of observations:** The observations are independent of each other.

```{r}
residuals <- fit$residuals
mean_res <- mean(residuals)
sd_res <- sd(residuals)
```

### 3. Normality Tests

Installing and loading the `"nortest"` package, which is used for a variety of normality tests, such as:

-   Kolmogorov-Smirnov Tests - `ks.test()`
-   Lilliefors Test - `lillie.test()`
-   Cramer-von Mises test - `cvm.test()`
-   Shapiro-Wilk Normality Test - `shapiro.test()`
-   Shapiro-Francia test - `sf.test()`
-   Anderson-Darling test - `ad.test()`

```{r warning=FALSE}
# install.packages(nortest)
library(nortest)

t1 <- ks.test(residuals, "pnorm", mean_res, sd_res)
t2 <- lillie.test(residuals)
t3 <- cvm.test(residuals)
t4 <- shapiro.test(residuals)
t5 <- sf.test(residuals)
t6 <- ad.test(residuals)
```

Taking the name, statistics, and p-values from each performed normality test and organizing them into a dataframe.

```{r}
tests <- c(t1$method, t2$method, t3$method, 
           t4$method, t5$method, t6$method)
statistics <- c(t1$statistic, t2$statistic, t3$statistic, 
                t4$statistic, t5$statistic, t6$statistic)
p_values <- c(t1$p.value, t2$p.value, t3$p.value, 
              t4$p.value, t5$p.value, t6$p.value)
```

```{r}
results <- data.frame(Test = tests, Statistic = statistics, P_Value = p_values)
knitr::kable(results)
```

### 4. Homoscedasticity Tests

```{r fig.height=6, fig.width=10}
boxplot(residuals ~ data$variety, 
        col="gray", ylab="Residuals", xlab="Varieties")
```

Installing and loading the package `lawstat` we can have acess to the function `levene.test()` to perform **Levene's Test of Equality of Variances**

```{r warning=FALSE}
# install.packages("lawstat")
library(lawstat)
levene.test(residuals, data$variety)
```

### 5. Independence Check

It\`s looks difficult to conclude something right?
```{r fig.height=5, fig.width=10}
# install.packages("car")
library(car)

qqPlot(residuals, 
       main = "Q-Q Plot with Confidence Envelope",
       col = "blue",
       pch = 20,
       envelope = 0.95)  # 95% confidence envelope
```

### 6. Multiple Comparisons

By installing and loading the package `ExpDes` we can speed up the process, utilizing `dic()` for and Completed Ramdomized Design (CMD)

```{r warning=FALSE}
# install.packages("ExpDes.pt")
library(ExpDes.pt)

dic(data$variety, data$yield, quali = TRUE, 
    mcomp = "lsdb", hvar= "bartlett", sigT = 0.05, sigF = 0.05)
```

### 7. CRD - Unbalanced Data

```{r}
remove(list=ls())
ls()
```

```{r}
data_unbalanced <- read.table("DataUnbalance.txt", head=TRUE)
data_unbalanced$vend <- factor(data_unbalanced$vend)

n <- length(data_unbalanced$result)

knitr::kable(data_unbalanced)
```

```{r}
fit_unbalanced <- aov(result ~ vend, data = data_unbalanced)
anova(fit_unbalanced)
```

```{r}
residuals_unbalanced <- fit_unbalanced$residuals
mean_res_unbalanced <- mean(residuals_unbalanced)
sd_res_unbalanced <- sd(residuals_unbalanced)
```

```{r fig.height=5, fig.width=10}
# install.packages("car")
library(car)

qqPlot(residuals_unbalanced, 
       main = "Q-Q Plot with Confidence Envelope",
       col = "blue",
       pch = 20,
       envelope = 0.95)  # 95% confidence envelope
```

```{r warning=FALSE}
library(nortest)
t1 <- ks.test(residuals_unbalanced, 
              "pnorm", mean_res_unbalanced, sd_res_unbalanced)
t2 <- lillie.test(residuals_unbalanced)
t3 <- cvm.test(residuals_unbalanced)
t4 <- shapiro.test(residuals_unbalanced)
t5 <- sf.test(residuals_unbalanced)
t6 <- ad.test(residuals_unbalanced)
```

```{r}
tests <- c(t1$method, t2$method, t3$method, t4$method, t5$method, t6$method)
statistics <- c(t1$statistic, 
                t2$statistic, 
                t3$statistic, 
                t4$statistic, 
                t5$statistic, 
                t6$statistic)

p_values <- c(t1$p.value, t2$p.value, t3$p.value, 
              t4$p.value, t5$p.value, t6$p.value)
```

```{r}
results <- data.frame(Test = tests, 
                      Statistic = statistics, 
                      P_Value = p_values)

knitr::kable(results)
```

```{r}
fit_final <- aov(result ~ vend, data = data_unbalanced)
anova(fit_final)
```

```{r warning=FALSE}
library(ExpDes.pt)
dic(data_unbalanced$vend, data_unbalanced$result, quali = TRUE, mcomp = "tukey", sigT = 0.05, sigF = 0.05)
```
