---
title: "Data transformation"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---

Reading data
```{r}
remove(list=ls())
data<-read.table("DataTransf.txt", head=TRUE)

data$trat<-factor(data$trat)

n <- length(data$n.est)

knitr::kable(head(data))
```
### 1. Metrics 

Calculating mean and variance
```{r}
mean.est <-tapply(data$n.est, data$trat, mean)
var.est <-tapply(data$n.est, data$trat, var)
```

Besides we are running an ANOVA model, it is just to get the residuals values
```{r}
fit1 <- aov(data$n.est~data$trat)
anova(fit1)

residuos <- data.frame(fit1$residuals)
res<-residuos[,1]

xb <- mean(res) # mu
sx <- sd(res) # sigma
```

### 2. ANOVA Presupsution

**2.1 Residual normality test**

```{r fig.height=6, fig.width=10, warning=FALSE}
library(car)

qqPlot(res, 
       main = "Q-Q Plot with Confidence Envelope",
       col = "blue",
       pch = 20,
       envelope = 0.95)  # 95% confidence envelope
```

```{r warning=FALSE}
library(nortest)
t1 <- ks.test(res, "pnorm", xb, sx);t1
t2 <- lillie.test(res);t2 
t3 <- cvm.test(res);t3 
t4 <- shapiro.test(res);t4 
t5 <- sf.test(res);t5 
t6 <- ad.test(res);t6 

tests <- c(t1$method, t2$method, t3$method, 
            t4$method, t5$method, t6$method)
st <- as.numeric(c(t1$statistic, t2$statistic, t3$statistic, 
                     t4$statistic, t5$statistic, t6$statistic))
pv <- c(t1$p.value, t2$p.value, t3$p.value, 
            t4$p.value, t5$p.value, t6$p.value)

results <- data.frame(Test = tests, Statistic = st, P_Value = pv)
knitr::kable(results)
```

**2.2. Homocedasticity test**

```{r}
bartlett.test(res, data$trat)
```
**2.3. Independence test**

```{r fig.height=6, fig.width=10}
plot(res,
     type = "b",
     pch = 20,
     col = "blue",
     xlab = "Observation Order",
     ylab = "Residuals",
     main = "Residuals in Order of Collection")
abline(h = 0, col = "red", lty = 2)
```
Does it say something?

###  3. Data transformation

**Which transformation should be used?**

When the statistical data comes from counts, it generally follows the *Poisson* distribution, where the mean and variance are equal. In this case, the square root transformation of the data stabilizes the variance, making it independent of the mean. This transformation can also be used when the variance is proportional to the mean.

```{r}
n_est_transf <- sqrt(data$n.est + 0.5)
data_tr <- data.frame(coleta = data$coleta, 
                      trat = data$trat, 
                      n_est_transf = n_est_transf); data_tr

knitr::kable(data_tr)
```

Calculation of means and variances of *n.est* transformed by treatment
```{r}
mean.est.transf <- tapply(n_est_transf, data$trat, mean); mean.est.transf
var.est.transf <- tapply(n_est_transf, data$trat, var); var.est.transf
```

Veryfing the residual variance from the transformated data
```{r}
fit2 <- aov(n_est_transf~data$trat)
anova(fit2)

res2 <- fit2$residuals; res2

xb <- mean(res2)
sx <- sd(res2) 
```

Error nomality verification
```{r}
nsim <- 500 # Number os observation
conf <- 0.95 # Conf of reability
```

We perform simulations to compare the observed residuals with the confidence envelope expected under the assumption of normality
```{r}
datasim <- matrix(rnorm(n * nsim, mean = xb, sd = sx), nrow = n)
datasim <- apply(datasim, 2, sort)

infsup <- apply(datasim, 1, quantile, 
                probs = c((1 - conf) / 2,(1 + conf) / 2))

xbsim <- rowMeans(datasim); xbsim
```
Residuals normality distribuition
```{r fig.height=6, fig.width=10}
library(car)

qqPlot(res2, 
       main = "Q-Q Plot with Confidence Envelope",
       col = "blue",
       pch = 20,
       envelope = 0.95) 
```

Verify using statistics numbers
```{r}
shapiro.test(res2)
```

Checking the homoscedasticity of residuals by Bartlett test
```{r}
bartlett.test(res2, data_tr$trat)
```

Checking the independence of the residual

The independence of the residuals can be assessed through a residuals vs. predicted values plot. If independence exists, there will be an absence of pattern.s

Residual vs. order of collection of the response variable values plot
data
```{r fig.height=6, fig.width=10}
# Se 'res2' é o vetor de resíduos e 'data$coleta' indica a ordem de coleta
plot(res2,
     type = "b",
     pch = 20,
     col = "blue",
     xlab = "Observation Order",
     ylab = "Residuals",
     main = "Residuals in Order of Collection")
abline(h = 0, col = "red", lty = 2)

```

```{r}
da <- data.frame(data_tr$coleta,res2)          
da1 <- da[order(data_tr$coleta, decreasing=F),]
```

```{r fig.height=6, fig.width=10}
plot(da1$res2,
     type = "b",
     pch = 20,
     col = "blue",
     xlab = "Observation Order",
     ylab = "Residuals",
     main = "Residuals in Order of Collection")
abline(h = 0, col = "red", lty = 2)
```

```{r}
fit3 <- aov(data_tr$n_est_transf ~ data_tr$trat)
summary(fit3)
```

```{r warning=FALSE}
require(ExpDes.pt)

fit4 <- dic(data_tr$trat, data_tr$n_est_trans, 
            quali = TRUE, mcomp = "snk", sigT = 0.05, sigF = 0.05)
```
