---
title: "Data transformation"
author: "Hernandes Peres"
date: "`r Sys.Date()`"
output: html_document
---

Exercicio de Transformacao

Um pesquisador realizou um experimento inteiramente casualizado com quatro variedades de pessegueiro, visando ao enraizamento de estacas. Após o tempo necessário, o pesquisador anotou o número de estacas enraizadas e obteve os resultados que se seguem.


```{r}

ls()# lista todos os objetos criados
remove(list=ls(all=TRUE)) #remove todos objetos criados
ls()

```

```{r}
dados<-read.table("DataTransf.txt", head=TRUE)
str(dados)

dados$trat<-factor(dados$trat)
str(dados)
dados
attach(dados)
str(dados)

n <- length(n.est)
cat("\n n =", n, "\n")

```

mean and variance determinate for each treat

```{r}
mean.est <-tapply(n.est, trat, mean,data=dados); mean.est 
var.est <-tapply(n.est, trat, var); var.est
```

res calculation

```{r}
fit1 <- aov(n.est~trat)
anova(fit1)
names(fit1)
fit1$fitted.values
residuos <- data.frame(fit1$residuals)
res<-residuos[,1]; res


xb <- mean(res) # mu
sx <- sd(res) # sigma
cat("\n Média amostral =", xb, "\n Desvio padrão amostral =", sx, "\n")

```

Analysis of residuals

```{r}
par(mfrow=c(2,2))    # Partição da tela gráfica
with(fit1,plot(res, col="blue", cex=0.7, xlab="Índices", ylab="Resíduos"))
abline(h=0,col="red", lwd=2)
with(fit1,qqnorm(res,col="blue",cex=0.7,xlab="Quantis teóricos",ylab="Quantis amostrais"))
with(fit1,qqline(res,col="red", lwd=2))
boxplot(res)
hist(rstandard(fit1),main="NULL",freq=F, ylim=c(0,0.6))
lines(seq(-1.5,3,0.01),dnorm(seq(-1.5,3,0.01)),col="red",lwd=3)
layout(1) # retorna a tela grafica para 1 linha e 1 coluna
```

residual normality

1- Gráfico QQ (Quantil-Quantil)

```{r}
qqn <- qqnorm(res,col="blue", cex=1, main = "", xlab = "Quantis teóricos N(0,1)", pch = 20,
ylab = "Quantis amostrais")
qqline(res, lty = 2, col = "red", lwd=2)

```

2- Gráfico QQ com envelope

```{r}
nsim <- 500 # Número de simulações
conf <- 0.95 # Coef. de confiança
# Dados simulados ~ normal
dadossim <- matrix(rnorm(n * nsim, mean = xb, sd = sx), nrow = n)
dadossim <- apply(dadossim, 2, sort)
# Limites da banda e média
infsup <- apply(dadossim, 1, quantile, probs = c((1 - conf) / 2,
(1 + conf) / 2))
xbsim <- rowMeans(dadossim); xbsim
# Gráfico
faixay <- range(res, dadossim)
qq0  <- qqnorm(res,col="blue", cex=1, main = "", xlab = "Quantis teóricos N(0,1)", pch = 20,
ylab = "Quantis dos Valores da Amostra", ylim = faixay)
eixox <- sort(qq0$x)
lines(eixox, xbsim,lty = 2, col = "red",lwd=2)
lines(eixox, infsup[1,])
lines(eixox, infsup[2,])
```

3- Normality test

```{r}
library(nortest)
t1 <- ks.test(res, "pnorm", xb, sx);t1 # Kolmogorov Smirnov
t2 <- lillie.test(res);t2 # Lilliefors
t3 <- cvm.test(res);t3 # Cramér-von Mises
t4 <- shapiro.test(res);t4 # Shapiro-Wilk
t5 <- sf.test(res);t5 # Shapiro-Francia
t6 <- ad.test(res);t6 # Anderson-Darling

# Tabela de resultados
testes <- c(t1$method, t2$method, t3$method, t4$method, t5$method,
t6$method)
estt <- as.numeric(c(t1$statistic, t2$statistic, t3$statistic,
t4$statistic, t5$statistic, t6$statistic))
valorp <- c(t1$p.value, t2$p.value, t3$p.value, t4$p.value, t5$p.value,
t6$p.value)
resultados <- cbind(estt, valorp)
rownames(resultados) <- testes
colnames(resultados) <- c("Estatística", "p-value")
print(resultados, digits = 4)

```

homocedasticity errors

```{r}

#1- Gafico diagnostico padrao para checar homocedasticidade.
plot(fitted(fit1), resid(fit1),pch=19, col=4, xlab="Valores ajustados (médias trat)", ylab="Resíduos")

#2- Gráfico dos resíduos vs tratamentos
plot.default(trat,res,pch=19, col=4, ylab="Resíduos",xlab="Tratamentos",)

##2- Teste de Bartlett (Pode haver número diferente de repetições por tratamento)
bartlett.test(res, trat)

```

# Data transformation

## Which transformation should be used?

When the statistical data comes from counts, it generally follows the Poisson distribution, where the mean and variance are equal. In this case, the square root transformation of the data stabilizes the variance, making it independent of the mean. This transformation can also be used when the variance is proportional to the mean.

```{r}
n_est_transf<-sqrt(n.est+0.5) # sqrt root
n_est_transf
dados_tr<-data.frame(coleta, trat, n_est_transf); dados_tr
str(dados_tr)

detach(dados) 
attach(dados_tr)
```

Calculation of means and variances of n.est transformed by trat

```{r}
mean.est.transf <-tapply(n_est_transf, trat, mean); mean.est.transf
var.est.transf <-tapply(n_est_transf, trat, var); var.est.transf
```

Anlasys of variance of the transformed data

```{r}
fit2<- aov(n_est_transf~trat)
anova(fit2)

res2 <- fit2$residuals; res2

# residuals estimations
xb <- mean(res2) # mean
sx <- sd(res2) # standard deviation
cat("\n Média amostral =", xb, "\n Desvio padrão amostral =", sx, "\n")
```

error nomality verification

```{r}
nsim <- 500 # Number os observation
conf <- 0.95 # Conf of reability
```

### Simulated data

```{r}
dadossim <- matrix(rnorm(n * nsim, mean = xb, sd = sx), nrow = n)
dadossim <- apply(dadossim, 2, sort)

infsup <- apply(dadossim, 1, quantile, probs = c((1 - conf) / 2,(1 + conf) / 2))
xbsim <- rowMeans(dadossim); xbsim
```

Chart

```{r}
faixay <- range(res2, dadossim)
qq0  <- qqnorm(res2,col="blue", cex=1, main = "", xlab = "Quantis teóricos N(0,1)", pch = 20,
ylab = "Quantis dos Valores da Amostra", ylim = faixay)
eixox <- sort(qq0$x)
lines(eixox, xbsim,lty = 2, col = "red",lwd=2)
lines(eixox, infsup[1,])
lines(eixox, infsup[2,])
```

Shapiro-Wilk test

```{r}
shapiro.test(res2)
```

Checking the homoscedasticity of the errors

```{r}
plot.default(trat,n_est_transf,pch=19, col=4, ylab="Número de Estacas Enraizadas",xlab="Tratamentos")
```

Bartlett test

```{r}
bartlett.test(res2, trat)
```

Checking the independence of the erros

The independence of the residuals can be assessed through a residuals vs. predicted values plot. If independence exists, there will be an absence of pattern.

```{r}
plot(fitted(fit2), resid(fit2),pch=19, col=4,xlab="Adjusted values (Transf data)", ylab="Res.")
```

Residual vs. order of collection of the response variable values plot

```{r}
da<-data.frame(coleta,res2);da             
da1<-da[order(coleta, decreasing=F),];da1 # Ordering
```

```{r}
plot(da1$coleta, da1$res2,pch=19,type = "o", col=2, xaxt="n",yaxt="n",
 xlab="Order of collected data", ylab="Res.")
axis(1, at = seq(0, 20, by = 1))
axis(2, at = seq(-1, 1, by = 0.1))
abline(h=0, col="blue" )

summary(res2)
abline(h=seq(-0.6,0.8,0.1),v=seq(1,20,1),lty=3,col="gray")
```

```{r}
fit3<- aov(n_est_transf~trat)
summary(fit3)
```

```{r}
require(ExpDes.pt)
fit4<- dic(dados_tr$trat, dados_tr$n_est_trans, quali = TRUE, mcomp = "snk", sigT = 0.05, sigF = 0.05)
```
